Environmental temperature is a strong driver of subspecies competition in the Drosophila microbiome

Version 1

Abstract

Most microbiome research focusses on the taxonomic composition at the species level to understand the impact of environmental factors, but intraspecific diversity has largely been ignored. To address this significant knowledge gap, we took advantage of the simple, culturable microbiome of Drosophila. First, we documented that natural populations of D. simulans harbor three diverged clades of Lactiplantibacillus plantarum, a key nutritional symbiont. We studied the distinct ecological roles of these three clades by exposing flies with their native microbiome to two temperature regimes in the laboratory. Tracking the three clades within the complete Drosophila microbiome over a period of more than 10 years at two temperatures, we identified strikingly distinct dynamics in response to the selection regime. We confirmed the functional differentiation of the three clades using in vitro growth dynamics and in vivo mono-association assays. Our results highlight that environmental selection operates at the subspecies level. Therefore, we conclude that the functional diversification of the microbiome can only be understood when intra– and interspecific diversity is considered.

Introduction

Intraspecific diversity has largely been overlooked in microbial ecology surveys. This can be in part attributed to the success of 16S rRNA profiling, which allows the characterization of species diversity with a very high efficiency (1,2), but cannot identify intraspecific diversity (3). Although environmental variation has been identified as one of the major factors driving species composition, its role for intraspecific adaptation has been largely overlooked (3). This blind spot of microbiome research is particularly surprising as it is known that conspecific strains, i.e. strains from the same species, can display large phenotypic variation (4). Clinically relevant species like Escherichia coli, in which different subspecies can be either pathogenic or commensal are a particularly illustrative example (5).

Although a few studies in environmental microbial communities have provided evidence for the functional importance of intraspecific variation (6,7), such analyses are considerably more difficult. Therefore, descriptive correlative studies are more common than functional assays (8). In the gut-dwelling bacterium B. fragilis conspecific strains compete for the same niche leading to the mutual exclusion of toxicogenic and non-toxicogenic strains in the gut environment (9). Not only in the human gut, but also in lakes, soil, and the sea conspecific strains frequently co-occur in sympatry (10–12). It has even been even reported that strain richness, i.e. sympatric strain diversity within the same species, stabilizes microbial communities in fluctuating environments (13,14).

In this work, we further explored the functional implications of intraspecific diversity within the same microbial population. We focused on the species Lactiplantibacillus plantarum, an extremely versatile lactic acid bacterium that has been associated with a variety of different habitats, including plants, the gastro-intestinal tracts of mammals and insects, as well as food such as meat, dairy, and pickled products (15). L. plantarum is a facultative symbiont of Drosophila (16–19). Drosophila larvae can enhance the growth of L. plantarum, which in response releases essential nutrients for the larvae under poor diet conditions (18). As a result of this nutritional symbiosis, L. plantarum is a prevalent taxon in the Drosophila microbiome (17).

To study the implications of intraspecific diversity of L. plantarum in fruit flies, we took advantage of an experimentally evolved Drosophila simulans population that has been adapting to two novel temperature regimes, hot and cold, for more than ten years. We detected three clades of L. plantarum co-occurring within the native microbiome of D. simulans population before the experimental temperature was modified. However, the exposure of the fly population to the novel temperature regimes altered the intraspecific composition of L. plantarum, revealing functional differences between the co-occurring clades. These functional differences were experimentally validated in vitro and in vivo. Our results provide a clear example for functional diversification of intraspecific diversity and demonstrate how experimental evolution can uncover this otherwise hidden functional diversity.

Results

Genome sequencing reveals the presence of three L. plantarum clades in the experimentally evolving populations

We sampled 33 isolates from the Florida experiment (See Materials and Methods for details) to characterize the diversity of L. plantarum (Figure 1, Table S1). The isolates formed three clades based on the pairwise average nucleotide identity (ANI). Between-group ANI values were 98.4-99.17%, below the ∼99.5% threshold for conspecific strain identity (20). On the other hand, the identity within groups was 99.48-99.99% (Figure 2). The clustering was consistent irrespective of whether we used SNPs in the core genome or presence/absence patterns of accessory genes. This indicates that the three clades are old with very limited genetic exchange.

Experimental setup to study the long-term evolution of L. plantarum in Drosophila simulans populations.

Pangenome of L. plantarum genomes isolated from the Florida experiment (Table S1), sorted by phylogeny.

Left panel: Maximum likelihood tree built based on the core genome alignment. Tree leaves correspond to the name of the isolate and are coloured based on isolation origin: ancestral population (grey), hot-evolved population (red) or cold-evolved population (blue). Numbers at nodes represent bootstrap support values based on 1000 replicates. Middle panel: Pattern of genes presence (black)/absence (white) in each genome. Right panel: pairwise average nucleotide identity between the isolates. The colour ranges between 98.5% (purple) and 100% (yellow).

Since the isolates were sampled either from the unevolved, hot-evolved or cold-evolved populations (Figure 1), we tested whether the clades were randomly distributed across the experimental treatments and observed a non-random distribution (Fisher’s exact test, 3×3 contingency table, p = 5.5e-10). In combination with the clustering of clades (Figure 2), we conclude that each of the three clades is more prevalent either in the novel experimental populations or in the unevolved founder population, at the beginning of the experiment. We therefore coined the terms H (hot-evolved), C (cold-evolved), and U (unevolved) clades, that will be used throughout the manuscript.

In order to further investigate the association between L. plantarum genotype and temperature, we sampled from two additional experiments, obtaining seven isolates from hot-evolved Drosophila simulans Portugal and two isolates from constant-cold-evolved Drosophila simulans South Africa (Table S1). In a phylogenetic tree, the hot-evolved isolates and the constant-cold-evolved isolates grouped into clades H and C, respectively (Figure S1). These results do not only support the association between L. plantarum clade composition and environmental temperature (Fisher’s exact test, 3×3 contingency table, p = 2.6e-12), but also suggest that clade C is favored irrespective of whether the temperature is fluctuating (20 °C during the day, 10 °C at night) or constant (15 °C). Furthermore, the grouping of isolates originating from different experiments to the same clades indicates that these are not restricted to a single natural D. simulans population, but represent global diversity.

The three sympatric clades independently adapted to their host

Our analyses suggest that all three clades are shared among natural D. simulans populations, but it is not clear whether they diverged in D. simulans or independently colonized their Drosophila host. We used 79 publicly available L. plantarum genomes from several environments and geographic locations (Table S2) to reconstruct the phylogeny based on the core genome.

All three L. plantarum clades are more closely-related to genomes from other sources than to each other, which implies that they diverged before colonizing D. simulans (Figure 3). Moreover, the genomes from the clade H are almost identical to public sequences isolated from global collections of Drosophila melanogaster (Figure 3, Table S2). The clade U isolate B246 is closely related to the Drosophila-associated isolate LpWF originating from a wild D. melanogaster individual (21). Clade C did not have a close relative in the collection. The high similarity of the L. plantarum sequences isolated from D. melanogaster and D. simulans highlights that both Drosophila species share this component of the microbiome. Most likely a deeper sampling of the D. melanogaster microbiome will also detect clade C.

Maximum likelihood tree of 92 L. plantarum genomes.

Tree leaves are coloured by source of isolation. The genomes from our lab are coloured in according to experimental treatment: unevolved in grey, hot-evolved in red, and cold-evolved in blue. Numbers at nodes represent bootstrap support values based on 1000 replicates.

Despite being phylogenetically divergent, the three sympatric clades could have converged functionally in the process of adaptation to a common environment. In order to test this, we built the L. plantarum pangenome including all the available isolates and performed hierarchical clustering based on the presence/absence of accessory genes. We found no grouping of the three sympatric clades (Figure S2), which suggests that they are not only phylogenetically diverged, but have the potential to be functionally different based on their pool of accessory genes.

Clade dynamics are temperature-specific and consistent between population replicates

Up to now, our analyses were restricted to isolates sampled at specific time points of the experiment, but since genomic DNA from flies with their microbiome was sequenced throughout the entire experiment (Figure 1), a metagenomic analysis could track the frequency trajectories of the three clades to shed more light on the adaptation process. We calculated the proportion of reads assigned to each clade in the reads from the population at a given time point. At the beginning of the experiment, the populations were dominated by clade U (54.5% relative abundance on average), followed by clade C (31.9%), and then clade H (13.6%) (Figure 4). However, when the fly populations were subjected to the experimental treatments, clade H became dominant under the hot regime and clade C became dominant under the cold regime. Clade U greatly decreased in relative abundance in both regimes. This suggests that changes in environmental temperature alter the fitness landscape of the population and lead to a new adaptive state, in which the relative fitness of the clades change.

Clade composition over time in each temperature regime depicted as the relative abundance of each L. plantarum clade (top: cold, bottom: hot).

Clade C in light blue, clade H in red and clade U in grey. For each temperature regime we summed the reads from the ten replicate populations.

A strong asset of the experimental evolution set up is the availability of 10 replicate populations that were generated from the same founder population and independently maintained throughout the entire experiment under the same conditions. Hence, the comparison of replicates provides an estimate for the strength of deterministic and stochastic forces during the experiment. In the cold regime, all ten replicates exhibited a rapid takeover of clade C (Figure S3). In contrast, the hot-evolved replicates were more variable in the rate at which clade H increased in abundance. The most extreme cases were replicates 8 and 10, in which clade C dominated the population for 90 generations before being replaced by clade H (Figure S3). Thus, the populations subjected to the hot regime took longer to reach a new equilibrium, which is also less reproducible than that of the cold regime. Overall, the temperature-specific dynamic is consistent across replicates in both regimes, confirming the association between environmental temperature and clade frequency that was observed during the isolation of the clones (Figures 1, 2 and S1).

Phenotypic differences in liquid culture

The consistent temperature-specific dynamics of L. plantarum in the replicate populations resulted in a different clade dominating each novel temperature regime. The fitness differences may arise from an increased growth rate at a given temperature. Alternatively, traits such as toxin-antitoxin systems or host interaction may cause temperature-specific fitness (22). To determine if the observed clade dynamics were driven by differences in temperature-dependent growth rates, we measured the growth rate of 29 isolates from the three clades at hot temperature conditions. These isolates represented all sampling events across the two temperature regimes (hot and cold) as well as the unevolved population (Table S1). The three groups of isolates differed significantly in growth rate, carrying capacity, and inflection time, i.e. the time it takes an isolate to reach the log phase (Figure 5; Kruskal-Wallis test, p < 0.001 for all parameters). Isolates belonging to clade C took on average more than 12 h longer to reach the log growth phase. Clade H had the highest growth rate, followed by U, and C. Clade U reached the highest density, followed by H, and C.

Clade-specific growth dynamics.

(A) L. plantarum growth curves overlapped. Lines are coloured by isolate’s clade. Background colour represents the growth temperature in the fluctuating environment. (B-D) Boxplots depicting the carrying capacity, growth rate, and inflection time of each clade. Each dot corresponds to a technical replicate. Data is grouped and coloured by clade. Statistical significance was determined using Dunn’s test with Holm-adjusted p-values. Only significant comparisons are indicated. **** p < 0.0001; *** p < 0.001; ** p < 0.01; * p < 0.05.

These results reveal the functional diversification of the three clades, but do not support the hypothesis that temperature-specific growth rates caused the shifts in composition. In liquid culture, clade U outperformed clade C in growth rate, carrying capacity, and inflection time despite the latter clade prevailed in the Drosophila populations for much longer (Figure 4, Figure S3). Nevertheless, we caution that the growth conditions in the Drosophila environment are radically different to the growth measured in liquid MRS (23). Furthermore, inside the fly gut L. plantarum lives in a complex community in which the different clades interact with each other, with the other components of the microbiome, and finally also with the host. Bacteria can persist in the community by thriving in the food or by stably colonizing the fly gut. The former strategy might be controlled by growth rate and carrying capacity, but the latter is affected by additional factors such as death rate in the gut and defecation rate, that cannot be assessed in liquid medium (21). Moreover, the resource availability is different in the solid fly food and in liquid MRS medium. Thus, observed differences in growth could indicate different strategies to persist in the community.

Fitness effects of L. plantarum strains on their host

Although growth dynamics differ significantly among the three clades in liquid culture, they do not explain the observed changes in abundance during experimental evolution. Therefore, we turned to the Drosophila host to explore the fitness consequences of colonization by each clade. L. plantarum has been shown to increase larval fitness of Drosophila melanogaster relative to germ-free flies (18,24,25). We hypothesized that the fitness advantage conferred by L. plantarum to its host could depend on the genotype of the symbiont and the environmental temperature, favoring a different clade in each regime. To test this, we designed a set of inoculation experiments in both focal temperatures using two host species, D. melanogaster and D. simulans, and the three L. plantarum clades. While we could produce axenic D. melanogaster, attempts to produce axenic D. simulans were unsuccessful. For this reason, inoculation experiments involving D. simulans were conducted using conventionally reared flies.

In axenic D. melanogaster none of the L. plantarum clades provided a fitness advantage to the host relative to germ-free controls, contrary to the effects reported in the literature. In the case of clades U and H, the number of offspring and developmental time did not differ significantly from the germ-free control (Figure 6 and S6; Dunn’s test, p > 1 for all pairs). This discrepancy could be explained by the supply of essential amino acids during larval growth by L. plantarum, (18,24–26). The high content of dried yeast 24.3 g/l in the fly food used in our experiment likely provided already sufficient amounts of essential amino acids, which negated the growth-promoting effects of L. plantarum. Even more surprising was the significantly reduced fitness in the presence of clade C in both temperature regimes (Figure 6; Dunn’s test, p < 0.05 for all significant comparisons). A mixture of all three clades resulted in a similar loss in fitness. The fitness reduction was stronger in the first transfer of flies, likely due to a higher bacterial load. This finding suggests an inverse relationship between bacterial abundance and host fitness.

Fitness effect of L. plantarum inoculation in axenic D. melanogaster.

(A, B) Total number of F1 flies eclosed normalised by day and female under the cold (A) and hot (B) regime. (C, D) Developmental time, estimated as the number of days it takes 50% of the offspring to eclose. Measurements were grouped by inoculation treatment and transfer. Each dot corresponds to a biological replicate (n = 10) Statistical significance was determined using Dunn’s test with Holm-adjusted p-values. Only significant comparisons are indicated. **** p < 0.0001; *** p < 0.001; ** p < 0.01; * p < 0.05.

In conventionally reared D. simulans we did not detect differences in the number of eclosed flies (Supplementary figure S6; ANOVA). However, the developmental time was significantly extended after inoculation with clade C at cold temperature (Dunn’s test, p < 0.05 05 for all significant comparisons). We attribute the weaker effects in the non-axenic flies to a reduced L. plantarum load due to the presence of other taxa (21), or to higher-order interactions with other members of the microbiome (27–29).

In summary, we observed that clade C, which is dominant in the cold-evolved populations, decreases host fitness when axenic flies are inoculated. The magnitude of this effect varies depending on the environmental temperature, the bacterial load, and the presence of other microbial taxa. Clade-specific effects of L. plantarum on the D. melanogaster host have been reported (19). In a low-protein diet, strains that were not isolated from Drosophila, enhanced larval growth relative to germ-free individuals, whereas another Drosophila-associated strain did not have any effect. Even a reduction of Drosophila lifespan has been reported in mono-association with L. plantarum (30), but the reduction of adult viability and developmental rate relative to the germ-free control is a novel finding in our study. We can conclude that the outcome of the L. plantarum-Drosophila symbiosis is clade-specific and can be detrimental to the host.

Functional divergence on the genomic level

Our analyses clearly indicate functional divergence between the three clades at genetic and phenotypic levels (see Figures S2 and 4-6). Therefore, we were interested in identifying the genes that contribute to this functional differentiation. A comparative KEGG pathway analysis shows that clade C contains the most KEGG Orthologs (KOs), 1163 KOs, followed by clade H (1157 KOs), and clade U (1131 KOs). A subset of 111 orthologs is unique to one or two clades. Of these, 41 were shared by clades H and C (Figure S7). Among these differential orthologs fell into 53 KEGG metabolic pathways.

Interestingly, we found a larger repertoire of sugar-related genes in the two lab-selected clades, C and H, than in clade U (Table S3, Figure S7). One of these genes encodes the enzyme hexosaminidase (EC: 3.2.1.52) and confers clades C and H the ability to use chitobiose as carbon source. This oligomer is the main product of chitin degradation, that makes up the flies’ peritrophic matrix and exoskeleton (31). Therefore, we speculate that the ability to exploit this ubiquitous source of carbon and nitrogen in the lab-maintained fruit flies, could be a strong target of selection in the lab environment. This could explain why clade U, which displayed a high growth rate and carrying capacity in liquid culture (Figure 5), is rapidly outcompeted in the fly populations (Figures 4 and S3).

Furthermore, clade H and C (Table S3) harbor different sugar-related functions, indicating that, despite their overlapping functions, these strains possess distinct sugar metabolic capacities. We also identified functional differences in cofactor biosynthesis and respiratory metabolism between clades. These differences do not explain the clade-specific selection, but reflect the different evolutionary histories of the clades (Table S3, figures S8-S11). A detailed description of the functional diversification can be found in the supplement.

Discussion

Intraspecific diversity within the same environment has been reported to be ecologically important to maintain the stability of the entire microbial community (11,13,14). This parameter has been studied in lakes, soils, seas, and the human gut (10–12). However, little attention has been given to intraspecific variation in Drosophila, and insect microbiomes in general (8). Based on our results in experimentally evolved fruit flies, we propose that within-species competition, thus far largely overlooked, could contribute to ecological adaptation and evolution of the host, as it has been observed at the species level in Drosophila (32) and other animals (33–35).

Our finding of three co-occurring subspecies of L. plantarum in our populations is particularly surprising, given the low microbiome richness of Drosophila at species level, which is estimated to be orders of magnitude lower than in humans (36,37). However, the intraspecific richness of L. plantarum in our flies was three times higher than that estimated in human gut microbiomes (11). This discrepancy between inter– and intra-species diversity suggests that the Drosophila microbiome is more complex than previously thought (16,27,38). However, this diversity exists below the species level and has therefore remained cryptic due to the limitations of amplicon sequencing (3).

By altering the experimental temperature at which the populations were maintained, we discovered consistent shifts in the composition of L. plantarum subspecies, that inevitably led to the dominance of a single clade. Further inoculation experiments revealed that clade composition significantly impacts the fitness of the host. L. plantarum genotypes differ in the fitness advantage that they provide to the host (19). However, to our knowledge, no L. plantarum genotype has yet been reported to negatively affect Drosophila development. Our finding shows that the well-characterized nutritional symbiosis between Drosophila and L. plantarum depends on the bacterial genotype and cannot be generalized to the entire species. Furthermore, clade C outcompeted the other two clades in the cold populations despite negatively affecting the host. This demonstrates that the benefits of parasitism or mutualism in the Drosophila microbiome are context-dependent, as already observed in humans (39).

Finally, we also gained insight into how the microbiome of the fly could have adapted to laboratory conditions. Unlike the non-selected clade U, the two lab-selected clades, C and H, had the unique genetic potential to exploit the byproducts of chitin degradation. Our results align with previous research on the “domestication” of the Drosophila microbiome. Comparative genomics in Acetobacter have revealed that bacteria associated with lab-reared Drosophila were selected for the presence of uricase genes and the loss of flagella, relative to their wild-isolated counterparts (40). In Acetobacter, the ability to degrade uric acid excreted by the host provides a significant competitive advantage (40). We hypothesize that the same may apply to the ability to metabolize host-derived chitin. We conclude that our findings are consistent with laboratory adaptation, and extends previous observations from Acetobacter to Lactiplantibacillus.

Overall, our work emphasizes the importance of intraspecific diversity in maintaining the stability of microbial communities and the host’s ability to adapt to environmental changes. The current climate crisis has sparked a significant interest in how environmental temperature affects microbial communities at the species level (41–43). Our study highlights that even subspecies diversity plays a key role in adaptation to environmental temperature, and probably other abiotic factors. Therefore, we propose that a comprehensive understanding of adaptive processes and ecological dynamics in the case of host-microbiome interactions depends critically on including diversity at the subspecies level.

Materials and Methods

Experimental evolution

Twenty replicate populations were set up using 202 isofemale lines from a natural D. simulans population collected in Florida (44). The populations were maintained in a 12 h photoperiod and at two different novel temperature regimes: ten replicates were kept in a fluctuating hot environment (28 °C during the day and 18 °C at night), whereas the other ten were kept in a fluctuating cold environment (20 °C during the day and 10 °C at night). The census population size of the replicates was 1000-1250 with a 50:50 sex ratio. The flies in each replicate were equally distributed across five 300 ml bottles containing 60 ml of standard Drosophila medium (300 g Agar + 990 g sugar beet syrup + 1000 g malt syrup + 2,310 g corn flour + 390 g soy flour + 900 g yeast in 37.5 L water) (45). The evolved D. simulans populations were sequenced in 10 generation intervals using Pool-Seq (46,47). These sequencing reads have been used to study the evolutionary dynamics of fruit flies (47). Similar pool-seq time series data have been previously used to analyze long-term endosymbiont and microbiome dynamics in D. melanogaster populations evolving in the same experimental conditions (48–50).

We also isolated L. plantarum from two other experimental evolution experiments, one originated from a South African population reared at constant 15 °C and the other from Portugal, which was reared in the fluctuating hot regime as described earlier (51). Both experimental evolution studies followed the same culturing regime that was used for the Florida populations, with temperature being the only experimentally altered variable (52).

Bacterial isolation

We primarily obtained data from the experimental evolution studies using the Florida D. simulans founder population and collected 33 L. plantarum genomes from five different time points; two time points from each experimental temperature and one time point corresponding to the ancestral flies (Table S1). In addition, we obtained L. plantarum genomes from two other experimental evolution studies. Two genomes from the South African experiment and seven genomes from the Portugal experiment.

Around 100 flies from each population were homogenized in sterile PBS using an autoclaved pestle. The homogenate was diluted to avoid physical contact between colonies and streaked on agar plates with three different media: De Man, Rogosa, and Sharpe (MRS), mannitol and tryptic soy. The plates were incubated at 28 °C for 48 h. We identified L. plantarum using PCR with custom Lactiplantibacillus-specific primers (in 5’-3’ orientation, LacF: GATGGGCGCTTACCCGATTA, LacR: CTGCCCCGCAAATTGTTTCA). The proportion of L. plantarum colonies in the plates, relative to other taxa, varied across growth media and fly populations. Colonies which could successfully be amplified were picked and regrown to stationary phase in the same medium from which they were isolated, but in liquid format. A glycerol stock of each isolate was made by mixing the culture with glycerol (50% v/v) in 1:1 proportion.

DNA extraction and sequencing

Genomic DNA was extracted from all the replicate populations using the high salt method (53). The fly populations were sampled in 10 generations intervals starting from generation 0. At sampling the age of the flies varied between four and eight days for the hot environment and between nine and 16 days for the cold environment. Pools of flies were sequenced at various time points, using a range of library kits, insert sizes, and read lengths (47).

For sequencing, the L. plantarum isolates were grown to stationary phase in MRS medium. gDNA was isolated with the high salt extraction method (53). Additionally, a lysozyme pre-treatment was used to degrade Gram-positive cell wall (54). Briefly, the pellet was resuspended in 480 µl of EDTA 50 mM. The suspension was treated with 120 µl of lysozyme (20 mg/ml dissolved in NaCl 30 mM – EDTA 2mM) and incubated for two hours at 37 °C. After this, the samples were centrifuged at maximum speed, the supernatant was discarded, and the pellet was further treated following the high salt gDNA extraction. Genomic DNA was quality-controlled on an Agilent Bioanalyzer (Agilent Technologies, Inc., Santa Clara, CA) and subsequently used to prepare DNBSEQ short-read libraries, and 2×150bp reads were sequenced on a DNBSEQ-G400 (MGI Tech Co., Ltd., Shenzhen, China).

Genomic assemblies

Sequencing reads were processed using BBDuk (BBMap, 2025 v38.90) to remove adapters, ΦX174 sequences and low-quality bases with the following parameters: ktrim=r, k=23, mink=11, hdist=1, tbo, tpe. Reads were assembled using SPAdes v4 (56) with kmer sizes 21, 33, 55, 77, 99, and 127. Subsequently, CheckM2 v1.0.1 (57) was used to estimate the completeness and contamination of each assembly, and the taxonomy was determined with GTDB-Tk v2.1.1 (58). Ten assemblies had more than 5% contamination, probably due to accidentally picking two adjacent colonies. We used Centrifuge v1.0.4 (59) to classify and exclude contigs with exogenous origin from these assemblies.

Comparative genomics and phylogeny

We built the pangenome and phylogenetic tree of the collection of genomes isolated from our lab (Table S1). First, we used Prokka v1.14.6 (60) to predict the genes in all the genomes. Roary v3.13 (61) was used to construct the isolates’ pangenome and extract the aligned genes belonging to the core genome, i.e. those that are present in all genomes. SNP-sites v2.5.1 (62) was used to extract the variable positions from the concatenated core genes. A maximum likelihood tree of the aforementioned variable positions was built using the model GTR+ASC and 1000 bootstraps with IQ-TREE v3.0.1 (63). Additionally, pairwise average nucleotide identity between all the isolates was computed using the program PyANI v0.2.13.1 (64), with ANIb as alignment method using the whole genomes. This workflow was carried out for the subset of 33 genomes isolated from the Florida experiment and for the whole dataset of 42 genomes, including those isolated from South Africa and Portugal experiments. Unless specified otherwise, subsequent analyses were conducted using the whole dataset pangenome.

The same phylogenetic and pangenomic methods were used for the extended L. plantarum genomes collection, including 79 publicly available genomes (Table S2). In order to root the tree, an additional tree was made including the reference genome from the sister species Lactiplantibacillus paraplantarum (GCA_003641145.1) as outgroup. The L. plantarum isolate that was most closely related to L. paraplantarum, NIZO2776, was used to root the tree.

In order to investigate which metabolic functions were differentially encoded in each clade, we followed the Anvi’o v7.1 pangenomic workflow (65,66) to identify “gene clusters” that are unique and ones that are shared across the genomes isolated from our lab. These gene clusters were further annotated against the KEGG database using the built-in Anvi’o function ‘anvi-run-kegg-kofams’. The script ‘anvi-compute-functional-enrichment’ (67) was then used to screen for KEGG functions that are differentially enriched between the clades. Briefly, this script estimates the fraction of genomes from each clade that encode each KEGG Ortholog (KO) and tests whether they are significantly associated to one or more groups. We used the R package ggKegg v1.4.1 (68) to visualize the KEGG pathways in which at least one gene was enriched.

Competitive mapping and clade relative abundance calculation

First, we pre-filtered the Drosophila pool-seq reads by removing the reads that have eukaryotic or Wolbachia origin. We used Bowtie v2.5.4 (69) with stringent settings (-D 500 –R 40 –N 0 –L 20 –I S,1,0.50 –-no-mixed –-no-discordant) to map each pool-seq data set against a collection of reference genomes that included D. simulans (GCF_016746395.2), D. melanogaster (GCF_000001215.4), D. mauritiana (GCF_004382145.1), H. sapiens (GRCh38), M. musculus (GCF_000001635.27), A. thaliana (GCF_000001735.4), S. cerevisiae (GCF_000146045.2), C. lupus (GCF_011100685.1) and a several insect-infecting Wolbachia strains (GCF_000008025.1, GCF_000022285.1, GCF_000376585.1, GCF_000376605.1, GCF_000475015.1). We kept the read pairs in which none of the reads mapped to any of the genomes in the collection, which should predominantly represent prokaryotic sequences.

Then, we used SuperPang v0.9.6 (70) to build the reference graph pangenome of each L. plantarum clade using all the genomes belonging to the clade as input (Table S1, Figure 3). In order to calculate the clades relative abundance in each pre-filtered pool-seq, we used BBMap (BBMap, 2025 v38.90) with “perfect mode” settings to competitively map each read against the three concatenated clade pangenomes. For each sample, the number of reads that mapped uniquely to each clade was counted and normalized to Reads Per Million (RPM) in order to account for differences in genome size and sample depth.

We benchmarked our approach to confirm the relative abundance estimates from our competitive mapping pipeline. Briefly, we simulated 200k HiSeq reads from each reference graph pangenome using InSilicoSeq v2.0.1 (71), and used ‘seqkit sampl’ v2.3.0 (72) to mix them in different ratios, maintaining the original size of 200k reads: 1:0:0, 0:1:0, 0:0:1, 1:1:1, 2:1:7, and 8:1:1. Each of the combined read sets was randomly subsampled from 150k to 10 reads (10, 20, 40, 60, 80, 100, 200, 300, 400, 500, 600, 700, 800, 900, 1k, 2k, 5k, 10k, 150k reads) and mapped competitively to the three clades. We confirmed that the relative abundances were correctly estimated with as few as 100-1000 reads uniquely mapping the references (Figure S4).

Growth assays

The glycerol stocks from all the available L. plantarum isolates were re-grown in agar MRS. A single colony per isolate was grown in liquid MRS and passaged daily for five days in order to remove the influence of freezing. On the fifth day, each suspension was normalized to an optical density (OD600) of 0.01, and serially diluted 1:100 in MRS. We selected 31 isolates that represented samples from all the available time points and temperature regimes, and plated them in triplicates in a flat-bottom 96-wells plate. Three wells were filled with sterile MRS medium as blank controls. Later, we detected contamination in two of the isolates, leaving us with 29 isolates. The plate was incubated for 72 h and OD600 was measured every 15 minutes in a Synergy H1 plate reader. The temperature regime was chosen to be similar to the hot conditions of the fly populations, but due to limitations of the spectrophotometer’s cooling capacity instead of 28 °C during the day and 18 °C at night, we set it to 28 °C during the day and 23 °C at night. Orbital shaking was set to 425 cpm. The R package GrowthCurver v0.3.1 (73) was used to infer the growth parameters from the growth curves. All statistical analyses were conducted in R v4.4.2 (74) using the package rstatix v0.7.2 (75). We used the Kruskal-Wallis test to assess overall differences in growth parameters among the clades. For post-hoc pairwise comparisons, Dunn’s test was applied, with p-value adjustments based on the Holm method. Differences were considered significant at p < 0.05.

Production of axenic flies

We generated germ-free flies, using a modified version of the dechorionation protocol described by Kietz et al. (76). Specifically, we used 2.8% active chlorine bleach instead of 1% and supplemented bleach, EtOH, and deionized H2O with TritonX (1% v/v). The addition of this detergent prevents dechorionated eggs from sticking to the walls of the tubes, thus facilitating their transfer to bottles. The whole process was carried out in sterile conditions using a laminar flow hood. Axenic flies were maintained via periodic transfers to new sterile food inside the laminar flow hood. We first carried out this procedure with several D. simulans isofemale lines, but it was not possible to fully eliminate the residual microbiome despite several rounds of treatment. However, we could produce and maintain axenic D. melanogaster Oregon-R flies. For this reason, we decided to use D. melanogaster as axenic host in the inoculation experiments.

Inoculation experiments

A representative isolate from each clade (Table S1, B89 for ancestral, S103 for cold, and S239 for hot) was re-grown from the glycerol stocks in MRS medium. After three transfers, OD600 was normalized to 0.05. Autoclaved vials with axenic fly food were inoculated with 50 µl of each bacterial suspension (10 vials per treatment). In total five treatments were used: the pure culture from each genotype, a mixture of the three genotypes in equal proportions (“mix”), and control, in which the vials were inoculated with 50 µl of fresh MRS medium. After overnight absorption of the liquid, twenty axenic flies (ten female and ten male flies) were added to each vial and allowed to lay eggs for either 24 h (hot conditions) or 48 h (cold conditions). After this step, the flies were transferred to a new set of vials with axenic, uninoculated food, and allowed to lay eggs for another 24 h (hot conditions) or 72 h (cold conditions). The adults were discarded and the vials were incubated at hot (28 °C during the day and 18 °C at night) or cold conditions (20 °C during the day and 10 °C at night) until the new generation eclosed. The number of eclosed flies was recorded daily. We estimated the total number of flies per vial and the developmental time as proxies of fitness. We calculated developmental time as the day from the start of the experiment until the day at which 50% of the flies in the vial were eclosed. The inoculation experiment was carried out with axenic D. melanogaster Oregon-R and non-axenic D. simulans from Florida. We used the Kruskal-Wallis test to assess overall differences in the two fitness phenotypes among the treatments. For post-hoc pairwise comparisons, Dunn’s test was applied, with p-values were adjusted using the Holm method. Differences were considered significant at p < 0.05.

eLife Assessment

This valuable study explores changes in the Drosophila microbiome in response to environmental temperature over more than ten years. The evidence showing that temperature leads to diversification of bacterial clades is solid, but additional information would help clarify how subspecies competition impacts microbiome composition and the host. The work will interest researchers working with microbiomes, microbial ecology, and evolutionary biology.

Reviewer #1 (Public review):

Summary:

The factors that create and maintain diversity in host-associated microbiomes remain poorly understood. A better understanding of these factors will help in the efforts to leverage the adaptive potential of the microbiome to help solve pressing problems in health and agriculture.

Experimental evolution provides a promising path forward as we can track the causes and consequences in the emergence of novel variants, but experimental evolution remains underutilized in host-microbiome interactions. Here, Gracia-Alvira utilizes a long-term experimental evolution study in Drosophila simulans under hot and cold temperature regimes to identify strain-level variation in an important fly bacterium, Lactiplantibacillus plantarum. They identify three strains of L. plantarum, which are most prevalent in their respective three temperature regimes, suggesting that these are locally adapted bacteria. Then, using a combination of genomics, in vitro, and in vivo, Gracia-Alvira et al attempt to understand the factors that led to the differentiation of the hot and cold L. plantarum and their impacts on the fly host.

Strengths:

This is an excellent use of experimental evolution to track the emergence of novelty in the microbiome. The genomic analyses are all solid and appropriate for the data sets. It is especially striking that the comparisons with the other, independent experimental evolution studies in different labs (and across continents between Portugal and South Africa) show a consistent response to temperature. Many have disregarded the microbiome as it is something that is too sensitive to seemingly innocuous variables (particularly in the fly microbiome), such that we cannot find generalities. However, this finding highlights the potential for experimental evolution to uncover these dynamics. The question of how strains emerge and are maintained is timely and is one of the key open questions in host-microbiome evolution currently.

Weaknesses:

(1) The framing in the title and throughout the discussion about "subspecies competition" does not match the data that was collected. The subspecies competition requires actually tracking the competitive outcomes between the hot, cold, and unevolved L. plantarum. In the in vivo work, I can see that mixes of the strains were made, but they did not track whether the cold strain outcompeted the hot strain in vivo under cold conditions, for example. While Figure 4 is suggestive that there is ongoing competition in the hot temperature regime, this is not necessarily shown in the cold, which is dominated by the C clade. It could also be that the bacteria cannot survive in the flies at the different temperatures. The growth curve assays hint that the bacteria can grow, but the plate reader couldn't actually maintain the 18 {degree sign}C temperature (line 455). So all of this evidence is very indirect and insufficient to say that strain competition is driving these patterns.

(2) The in vivo results are interesting in that there appears to be a fitness cost of clade C, but the explanation is underdeveloped. I say under-developed because in Figure 4, the cold L. plantarum remains much higher throughout adaptation to the hot temperature regime than the hot L. plantarum in the cold regime. The hot L. plantarum is low abundance throughout the cold regime. I felt like this observation was not explained, but it seems relevant to understanding the strain dynamics.

I will also note that this is not the first time that L. plantarum or other Lactobacillus have been shown to exert fitness costs to Drosophila. Gould, PNAS, 2018, shows that both Lactobacillus plantarum and Lactobacillus brevis in mono-association have lower fitness (measured through Leslie matrix projections using lifespan and fecundity) than axenic flies. Many studies of wild Drosophila fail to find Lactobacillus, or it is low abundance (e.g., Chandler, PLoS Genetics, 2014; Wang, Environmental Microbiology Reports, 2018; Henry & Ayroles, Molecular Ecology, 2022; Gale, AEM, 2025). This might help provide useful context for the in vivo results.

(3) The data in Figure 4 are compelling to focus on the L. plantarum variants. However, I can see from the methods that the competitive mapping included only other strains of Wolbachia. It is not clear how other members of the microbiome changed in response to the temperature regimes. As I note in point #2, given that Lactobacillus is often rare, it is not clear what the rest of the microbiome looks like over the course of adaptation. Indeed, it seems like Mazzucco & Schlotterer, PRSB, 2021 did a broader analysis of the microbiome and found that Acetobacter is by far the most common bacterium (I think this data is also part of the data shown here?). Expanding on why or why not in this context is important and will improve this study, particularly if the focus is on connecting these evolutionary dynamics to ecological competition to explain the emergence of strain diversity.

Reviewer #2 (Public review):

Summary:

In this manuscript, Gracia-Alvira et al. investigated how environmental temperature affects competition among members of the microbiome, with a focus on intraspecific diversity, using the Drosophila model.

Notably, the authors identified three clades of Lactiplantibacillus plantarum from a natural population of Drosophila simulans collected in Florida. They tracked the dynamics of these three bacterial clades under two temperature conditions over the course of more than ten years. Using comparative genomics and phylogeny, they showed that these three bacterial clades likely adapted to their host independently in a temperature-specific manner. Further, by combining in vitro culture and in vivo mono-association assays, they demonstrated the functional divergence of these three bacterial clades phenotypically, including their growth dynamics and effects on host fitness. Lastly, they performed pathway analysis and speculated on key genomic variance supporting such functional divergence.

Strengths:

The laboratory evolutionary experiment in response to cold or hot environmental temperature is impressive, given its more than ten years of experimental time period. This collection of achieved microbiome samples paired with the fly host data can be a valuable resource for the field.

Weaknesses:

The laboratory evolutionary experiment can be limited due to its artificial experimental setup. For example, wild flies rely on a more diverse set of food sources and are constantly exposed to new bacterial inoculations, whereas under laboratory conditions, flies live in a more restricted ecosystem. In addition, environmental temperatures differ among different locations, but they also involve seasonal changes within the same region. This manuscript can be strengthened with further discussions that elaborate on these limitations.

Moreover, the extent of host effects involved in these experiments remains ambiguous, because it is unclear whether these Lactiplantibacillus plantarum mostly reside within fly guts or on Drosophila medium. The laboratory evolutionary experiment possibly favored better colonizers on Drosophila medium under either cold or hot temperatures, which subsequently can saturate fly guts. As fully dissociating these variables can be experimentally tedious, the authors may want to comment more on these aspects in the discussion. Or they may want to consider some measurements. For example, measuring the growth rate of these bacteria on Drosophila medium under different temperatures, in addition to the current MRS culture experiments, or measuring the portion of the Lactiplantibacillus on Drosophila medium versus these stably colonizing fly guts.

Reviewer #3 (Public review):

Summary:

The study presents an analysis of 297 pangenomes derived from 20 populations of Drosophila simulans, at 19 time points for fast-reproducing individuals in a hot environment, or at 10 time points for slow-reproducing individuals in a cold environment, over a period of more than 10 years. The authors select a particular microbial component of the pangenomes and study the dynamics of Lactiplantibacillus plantarum strains in two environments. They discover that the revealed operational taxonomic units could be divided into three phylogenetic clades, which have their own genomic and genetic features, different adaptive capabilities that depend on the environment, and have a distinct impact on the fitness of the host.

Strengths:

The authors prove that bacterial microbiome components are sensitive to the environment and could rapidly (years) be fixed in eukaryotic populations. This study establishes a tractable model that potentially enables the study of variability of the physiological influence of distinct strains of an important commensal species, Lactiplantibacillus plantarum, on the Drsosophila host. It is clearly shown that this single species consists of several phylogenetically and functionally diverse strains. The authors did not limit their interest to their own model, but rather they have integrated a comparative approach by analysing phylogenetic relationships among 92 described L.plantarum strains.

Overall, the study is novel and delivers important discoveries of a longitudinal, well-replicated experiment, generating a substantial amount of genomic data. It highlights an important dimension of research that environmental selection operates at the subspecies level.

Weaknesses:

Even though the authors show only one particular example by conducting their longitudinal experiment, they honestly acknowledge failures important for interpretation of the biological significance of the results (gnotobiotic mono-association experiments was done with D.melanogaster, but not D. simulans) and therefore they state limitations of their conclusions (weaker effects in the non-axenic flies are due to the presence of other taxa or to higher-order interactions with other members of the microbiome). These interactions could significantly affect bacterial growth, metabolism, and physiological influence on the host.

The authors exploit the results of their experiment to speculate about a wide range of evolutionary phenomena, like within-species competition, ecological adaptation and evolution of the host, fitness advantage of bacteria to the host, the benefits of parasitism or mutualism, the domestication of the microbiome, etc. At the end, they conclude that their study "highlights that even subspecies diversity plays a key role in adaptation to environmental temperature". However, the potential mechanisms of such adaptation are barely discussed, so that the focus of the study shifts from the temperature-induced changes in microbial population structures toward metabolism-related adaptations of clade representatives that enable them to diversify their carbon and nitrogen sources. The role of the temperature factor remains elusive.

In addition to that, the paper has a clearly minimalistic experimental approach to address functional properties of the revealed L.plantarum strains, so that their own fitness, or their relationship with the Drosophila host, is characterised superficially. Therefore, the authors' discourse can be speculative rather than factual (especially when the authors use the expression "likely" to share their guesses in the "Results" section). Nevertheless, these minor drawbacks do not underscore the novelty of the discovered phenotypes and the importance of their further investigation.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

The factors that create and maintain diversity in host-associated microbiomes remain poorly understood. A better understanding of these factors will help in the efforts to leverage the adaptive potential of the microbiome to help solve pressing problems in health and agriculture.

Experimental evolution provides a promising path forward as we can track the causes and consequences in the emergence of novel variants, but experimental evolution remains underutilized in host-microbiome interactions. Here, Gracia-Alvira utilizes a long-term experimental evolution study in Drosophila simulans under hot and cold temperature regimes to identify strain-level variation in an important fly bacterium, Lactiplantibacillus plantarum. They identify three strains of L. plantarum, which are most prevalent in their respective three temperature regimes, suggesting that these are locally adapted bacteria. Then, using a combination of genomics, in vitro, and in vivo, Gracia-Alvira et al attempt to understand the factors that led to the differentiation of the hot and cold L. plantarum and their impacts on the fly host.

Strengths:

This is an excellent use of experimental evolution to track the emergence of novelty in the microbiome. The genomic analyses are all solid and appropriate for the data sets. It is especially striking that the comparisons with the other, independent experimental evolution studies in different labs (and across continents between Portugal and South Africa) show a consistent response to temperature. Many have disregarded the microbiome as it is something that is too sensitive to seemingly innocuous variables (particularly in the fly microbiome), such that we cannot find generalities. However, this finding highlights the potential for experimental evolution to uncover these dynamics. The question of how strains emerge and are maintained is timely and is one of the key open questions in host-microbiome evolution currently.

Weaknesses:

(1) The framing in the title and throughout the discussion about "subspecies competition" does not match the data that was collected. The subspecies competition requires actually tracking the competitive outcomes between the hot, cold, and unevolved L. plantarum. In the in vivo work, I can see that mixes of the strains were made, but they did not track whether the cold strain outcompeted the hot strain in vivo under cold conditions, for example.

We thank the reviewer for the honest concern and take this opportunity to defend our claim of "subspecies competition used across the manuscript. As the reviewer states, subspecies competition requires tracking the competitive outcomes between the three clades, and this is what we did by sampling and sequencing across ten years of experimental evolution (Figures 4 and S3). For this reason, we point that the subspecies competition assessment comes from the direct observation of changes in relative abundance across the time series, and not from the follow-up experiments in vivo or in vitro.

While Figure 4 is suggestive that there is ongoing competition in the hot temperature regime, this is not necessarily shown in the cold, which is dominated by the C clade. It could also be that the bacteria cannot survive in the flies at the different temperatures. The growth curve assays hint that the bacteria can grow, but the plate reader couldn't actually maintain the 18 {degree sign}C temperature (line 455). So all of this evidence is very indirect and insufficient to say that strain competition is driving these patterns.

We thank the reviewer for the alternative hypothesis that could explain the observed subspecies dynamic. We rule out that dominance of clade C in the cold occurs because the other two clades cannot grow in this regime based on three pieces of evidence:

(1) In the time series, clades H and U decrease, but never disappear (Figures 4 and S3), even showing some peaks of abundance in specific replicate populations (Figure S3).

(2) We isolated individuals belonging to clade H in the cold-evolved populations, as shown in figure 2. This is a direct evidence that clade H prevails in the cold-evolved populations, although in low abundance.

(3) We did grow the three taxa in fly food Petri dishes incubated at both temperature regimes, observing growth in all cases.

We will include the food growth experiment in the revised manuscript as further supporting evidence for growth in both regimes.

(2) The in vivo results are interesting in that there appears to be a fitness cost of clade C, but the explanation is underdeveloped. I say under-developed because in Figure 4, the cold L. plantarum remains much higher throughout adaptation to the hot temperature regime than the hot L. plantarum in the cold regime. The hot L. plantarum is low abundance throughout the cold regime. I felt like this observation was not explained, but it seems relevant to understanding the strain dynamics.

We acknowledge that a strong fitness cost of clade C is observed in axenic D. melanogaster. In the native host, D. simulans, with reduced microbiome, we observed delayed development that could even be an advantage depending on the situation, as pointed out by reviewer 3 in the recommendations.

Even if we assume that flies colonized with clade C are less fit in the experimental evolution, another caveat is whether the flies can actively select for the L. plantarum clade. Under this assumption, a clade that imposes a fitness cost to the fly (clade C) should be selected against over time because the flies colonized by this clade will have less offspring or develop later than the rest. Alternatively, as the microbiome is shared among all the individuals in the population, the host might not be able to “purge” the pernicious clade, and L. plantarum dynamics might be controlled solely by the relative fitness between clades in the given experimental treatment. We will discuss this hypothesis in the revision as a way to explain the relationship between the abundance of each clade and the effect on the host.

I will also note that this is not the first time that L. plantarum or other Lactobacillus have been shown to exert fitness costs to Drosophila. Gould, PNAS, 2018, shows that both Lactobacillus plantarum and Lactobacillus brevis in mono-association have lower fitness (measured through Leslie matrix projections using lifespan and fecundity) than axenic flies. Many studies of wild Drosophila fail to find Lactobacillus, or it is low abundance (e.g., Chandler, PLoS Genetics, 2014; Wang, Environmental Microbiology Reports, 2018; Henry & Ayroles, Molecular Ecology, 2022; Gale, AEM, 2025). This might help provide useful context for the in vivo results.

We thank the reviewer for the references. These observations are compared to our phenotypic results and discussed in the revised version of the manuscript.

(3) The data in Figure 4 are compelling to focus on the L. plantarum variants. However, I can see from the methods that the competitive mapping included only other strains of Wolbachia.

We appreciate the thorough reading of the methods by the reviewer. The competitive mapping comprised two steps: first we discarded the reads that mapped to Drosophila, Wolbachia and additional potential contaminants from sequencing facitilies (human, dog...). This step leaves the reads originated from whole the external microbiome of the flies, including L. plantarum. The second competitive mapping step recruits the reads that map any clade of L. plantarum.

It is not clear how other members of the microbiome changed in response to the temperature regimes. As I note in point #2, given that Lactobacillus is often rare, it is not clear what the rest of the microbiome looks like over the course of adaptation. Indeed, it seems like Mazzucco & Schlotterer, PRSB, 2021 did a broader analysis of the microbiome and found that Acetobacter is by far the most common bacterium (I think this data is also part of the data shown here?). Expanding on why or why not in this context is important and will improve this study, particularly if the focus is on connecting these evolutionary dynamics to ecological competition to explain the emergence of strain diversity.

We acknowledge that the rest of the Drosophila microbiome is not addressed in this study, as we wanted to focus the storyline around the intraspecific dynamics found in L. plantarum. We consider that a complete characterization of the whole Drosophila microbiome would unnecessarily elongate the paper and thus we treat it as a constant biotic factor.

We must point out that our dataset is not the one reported by Mazzucco & Schlötterer, which was done in D. melanogaster, rather than D. simulans. Nevertheless, both experiments share the same infrastructure, temperature regimes and fly maintenance.

We have included a list of taxa that were isolated from the populations, as well as to report L. plantarum prevalence and abundance across the experiment in order to provide context of the microbiome, beyond L. plantarum, to the readership.

Reviewer #2 (Public review):

Summary:

In this manuscript, Gracia-Alvira et al. investigated how environmental temperature affects competition among members of the microbiome, with a focus on intraspecific diversity, using the Drosophila model.

Notably, the authors identified three clades of Lactiplantibacillus plantarum from a natural population of Drosophila simulans collected in Florida. They tracked the dynamics of these three bacterial clades under two temperature conditions over the course of more than ten years. Using comparative genomics and phylogeny, they showed that these three bacterial clades likely adapted to their host independently in a temperature-specific manner. Further, by combining in vitro culture and in vivo mono-association assays, they demonstrated the functional divergence of these three bacterial clades phenotypically, including their growth dynamics and effects on host fitness. Lastly, they performed pathway analysis and speculated on key genomic variance supporting such functional divergence.

Strengths:

The laboratory evolutionary experiment in response to cold or hot environmental temperature is impressive, given its more than ten years of experimental time period. This collection of achieved microbiome samples paired with the fly host data can be a valuable resource for the field.

Weaknesses:

The laboratory evolutionary experiment can be limited due to its artificial experimental setup. For example, wild flies rely on a more diverse set of food sources and are constantly exposed to new bacterial inoculations, whereas under laboratory conditions, flies live in a more restricted ecosystem. In addition, environmental temperatures differ among different locations, but they also involve seasonal changes within the same region. This manuscript can be strengthened with further discussions that elaborate on these limitations.

As the reviewer has correctly noted, our experimental setting is not exempt from limitations. Lab-reared flies are fed with a defined standard diet. Furthermore, although the system is not completely closed to bacterial migration, this is limited as replicate populations are not allowed to mix during the maintenance of the flies. For this reason, we consider our laboratory setting as a compromise between observing wild populations, which undergo all biotic and abiotic stresses but cannot be manipulated, and evolving the bacteria in absence of the host, or in gnobiotic hosts, in which biotic interactions are not fully considered. We will extend on this in the new version of the manuscript.

Moreover, the extent of host effects involved in these experiments remains ambiguous, because it is unclear whether these Lactiplantibacillus plantarum mostly reside within fly guts or on Drosophila medium. The laboratory evolutionary experiment possibly favored better colonizers on Drosophila medium under either cold or hot temperatures, which subsequently can saturate fly guts. As fully dissociating these variables can be experimentally tedious, the authors may want to comment more on these aspects in the discussion. Or they may want to consider some measurements. For example, measuring the growth rate of these bacteria on Drosophila medium under different temperatures, in addition to the current MRS culture experiments, or measuring the portion of the Lactiplantibacillus on Drosophila medium versus these stably colonizing fly guts.

The reviewer's point was briefly addressed in the Results chapter: "Phenotypic differences in liquid culture".

Reviewer #3 (Public review):

Summary:

The study presents an analysis of 297 pangenomes derived from 20 populations of Drosophila simulans, at 19 time points for fast-reproducing individuals in a hot environment, or at 10 time points for slow-reproducing individuals in a cold environment, over a period of more than 10 years. The authors select a particular microbial component of the pangenomes and study the dynamics of Lactiplantibacillus plantarum strains in two environments. They discover that the revealed operational taxonomic units could be divided into three phylogenetic clades, which have their own genomic and genetic features, different adaptive capabilities that depend on the environment, and have a distinct impact on the fitness of the host.

Strengths:

The authors prove that bacterial microbiome components are sensitive to the environment and could rapidly (years) be fixed in eukaryotic populations. This study establishes a tractable model that potentially enables the study of variability of the physiological influence of distinct strains of an important commensal species, Lactiplantibacillus plantarum, on the Drosophila host. It is clearly shown that this single species consists of several phylogenetically and functionally diverse strains. The authors did not limit their interest to their own model, but rather they have integrated a comparative approach by analysing phylogenetic relationships among 92 described L. plantarum strains.

Overall, the study is novel and delivers important discoveries of a longitudinal, well replicated experiment, generating a substantial amount of genomic data. It highlights an important dimension of research that environmental selection operates at the subspecies level.

Weaknesses:

Even though the authors show only one particular example by conducting their longitudinal experiment, they honestly acknowledge failures important for interpretation of the biological significance of the results (gnotobiotic mono-association experiments was done with D. melanogaster, but not D. simulans) and therefore they state limitations of their conclusions (weaker effects in the non-axenic flies are due to the presence of other taxa or to higher-order interactions with other members of the microbiome). These interactions could significantly affect bacterial growth, metabolism, and physiological influence on the host.

We agree with the reviewer in that the use gnobiotic animals is a limitation, as by "tuning" the flies' microbiome we are modifying the interactions between members, which can potentially change the phenotypic outcome. Nevertheless, we use it as a complementary approach, rather than the only inference in our study.

The authors exploit the results of their experiment to speculate about a wide range of evolutionary phenomena, like within-species competition, ecological adaptation and evolution of the host, fitness advantage of bacteria to the host, the benefits of parasitism or mutualism, the domestication of the microbiome, etc. At the end, they conclude that their study "highlights that even subspecies diversity plays a key role in adaptation to environmental temperature". However, the potential mechanisms of such adaptation are barely discussed, so that the focus of the study shifts from the temperature-induced changes in microbial population structures toward metabolism-related adaptations of clade representatives that enable them to diversify their carbon and nitrogen sources. The role of the temperature factor remains elusive.

We acknowledge that our study does not fully resolve the mechanism by which a different clade ends up dominating each temperature regime. The MRS liquid experiment was an attempt to answer whether differences in optimal growth temperature could explain the temperature-specific abundance of the two clades. Our experiments showed, however, that this was not the case. Beyond this point, it is hard to disentangle the role of the temperature, as it could also act indirectly on the bacteria, for example, through the host or the food.

A second observation in our time series was that a third clade, U, was unfit in both regimes despite starting the experiment in high abundance. For this reason we also studied what made this clade less fit. Based on our analyses, we propose that the decrease of clade U was driven by the shift to a laboratory diet, shared by all experimental populations.

In addition to that, the paper has a clearly minimalistic experimental approach to address functional properties of the revealed L. plantarum strains, so that their own fitness, or their relationship with the Drosophila host, is characterised superficially. Therefore, the authors' discourse can be speculative rather than factual (especially when the authors use the expression "likely" to share their guesses in the "Results" section). Nevertheless, these minor drawbacks do not underscore the novelty of the discovered phenotypes and the importance of their further investigation.

We consider the reviewer's concern and toned down the phrasing when reporting our findings in the revised version of the manuscript.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

(1) One solution to resolve the "competition" issue would be to check that the L. plantarum strains are established at similar or different titers in the in vivo work. Fly phenotypes can be sensitive to microbial load (Keebaugh, iScience, 2018), which might explain some of the counterintuitive in vivo results. In line 227, the authors mention that "bacterial load" is contributing to the magnitude of the effect, but I don't see the data reported anywhere. If this is from the in vitro assays, then the authors need to show that in vitro predicts in vivo L. plantarum abundance.

Bacterial load inoculated in the in vivo experiment was normalized to OD=0.05 (~5*106 CFUs/ml) for the three clades at the beginning of the experiment. Thus, all vials were inoculated with the same titer of L. plantarum. Only the genotype varied between treatments. However, it is possible that, once inoculated, each clade grew to a different titer (as they have different growth rate and carrying capacity).

Statement in line 227 comes from the differing results in transfers 1 and 2. In transfer 1 we inoculated a fixed load of ~2.5*105 CFUs. In transfer 2, however, we inoculated no bacteria to the food, and the flies seeded the vial. Our statement comes from the assumption that bacterial load in transfer 2 has to be lower than in transfer 1 as bacteria seed the vial solely by defecation of the parents.

Following the reviewer's suggestion, in the revised version of the manuscript we have included a new experiment in which we quantified the bacterial load of each clade in individual flies.

(2) Tracking the competitive outcomes is tricky, though it could be done with whole genome sequencing. An alternative would be to label the strains with fluorescent proteins (e.g., Obadia Current Biology 2018 has done this in Lactobacillus) and track fluorescence to better understand the results of the "mix" treatment in Figure 6.

We appreciate the feedback of the reviewer, but consider this rather labour-intensive approach as an interesting option for future follow-up experiments.

(3) That being said, my main concern with this is the "competition" claim. If the paper were reframed appropriately, this paper could still make an important contribution to the evolution of host-microbiome interactions, but the authors would need to consider what they can and cannot do with this interesting dataset.

The "competition" claim comes from the changes in relative abundance observed in the time-series data, not from any of the follow-up experiments. Thus, we consider the use of the term "competition" appropriate.

(4) The text on the figures is very small and hard to read.

We increased the size of the text in all figures.

Reviewer #2 (Recommendations for the authors):

(1) Have you conducted the in vitro culture experiments following the "cold" conditions?

We have conducted the experiment in "cold" conditions, but with some modifications to the experimental settings, as the plate reader did not have cooling capacity. Instead, we grew a subset of the isolates (four per clade) in glass vials at constant 20 °C, and measured their OD twice a day. We have included the results in the revised version of the manuscript.

(2) How many technical and biological replicates were measured for the in vitro culture experiments (Figure 5)? Please add this information to the figure legend and method.

We measured the growth of four isolates from clade U, nine isolates from clade C and sixteen isolates from clade H. Each isolate was grown three times.

We have included this information, as requested by the reviewer.

(3) Making the labels in Figures 2, 3, 5, and 6 bigger would be helpful.

We have increased the font size of all figures.

Reviewer #3 (Recommendations for the authors):

(1) Line 268: "Based on our results in experimentally evolved fruit flies, we propose that within-species competition, thus far largely overlooked, could contribute to ecological adaptation and evolution of the host". Overstatement should be avoided, since the evolution of the host was not directly studied here.

Our results show that reproductive traits of the host differ upon colonization with each clade. Although we don't test the host's evolution, we speculate that flies differing in their offspring number and developmental time might differ in their overall fitness. Finally, we consider the Discussion section as the right place for speculation and development of hypotheses that can be tested in future work.

(2) Line 258: "These differences do not explain the clade-specific selection, but reflect the different evolutionary histories of the clades". The temperature factor and its possible role in clade selection would be better discussed at least a little bit.

In this paragraph we described potential metabolic differences between clades using comparative genomics. We did not find enrichment in a function or group of functions that could explain the different dynamics between clades H and C in the temperature regime.

In the revised version of the manuscript we highlight that we did not find temperature-specific differences from this analysis.

(3) Line 252: "...This could explain why clade U, which displayed a high growth rate and carrying capacity in liquid culture". The statement could be further developed with a caution. Even if the isolates that belong to the clade U are outcompeted by H or C, it should be noted that the strain U cannot be used as a true reference for fitness, since it could possess its hidden adaptive properties, not being simply "a loser". Such a hypothesis could explain the maintenance of this strain in the wild.

We agree with the reviewer in that fitness is relative to the selective environment. Clade U is less fit than H and C in our specific experimental conditions, but it could outcompete them in other conditions, such as wild flies or MRS liquid medium. In the revised version of the manuscript we have rephrased this statement to clarify that we specifically refer to clade U's fitness under the new laboratory conditions.

(4) In a cold environment, association with the clade C induces developmental delay and produces less progeny, which potentially allows the host to survive in case of harsh conditions and potential food limitation. Could the authors speculate and not exclude that this phenotype could be potentially adaptive? It would be curious to check in further studies whether flies associated with C strains are more stress-resistant, for example.

We thank the reviewer for this alternative hypothesis. In our manuscript we used the Darwinian definition of fitness; reproductive success of an organism in the focal environment. And thus, both higher progeny per female and shorter developmental time would be beneficial in direct competition with other individuals. It is true that delayed developmental time, or less progeny, could be advantageous in specific cases. This could be the case for D. simulans inoculated with clade C. However, we consider that the fecundity levels observed in D. melanogaster upon inoculation with Clade C (average of 0.06 offspring/female*day in the cold) are too low to sustain a population.

We have included this hypothesis in the Results section.

(5) It would be highly recommended to add an experiment to complete the story by measuring the quantity of bacteria in the medium and in the flies. This will resolve the hypothesis (Line 785): "Thus, the ability to exploit this ubiquitous source of carbon and nitrogen could be very advantageous in the fly microbiome context, but would not affect the fitness in liquid culture".

Following the reviewer's recommendation we included two additional experiments. We measured the bacterial load per fly in the native host, D. simulans, inoculated with the three clades. We also compared the clades' growth speed in solid fly food (without host). In the former experiment, we found similar bacterial loads upon inoculation with clades U and clade H. In contrast, in the latter we found delayed growth of clade U relative to H and C in the food. Thus, chitobiose consumption does not seem to provide an advantage in the fly gut to clade H. We attribute the fitness advantage of H and C to their advantage growing on the laboratory fly food, regardless of the host.

Both experimental results have been included in the revised version of the manuscript, and the comparative genomics paragraph and discussion have been modified in consequence.

(6) The chapter "Extended clade-specific differences in KEGG metabolic pathways" could be presented in the main text as it contains important results. These results are mentioned in the chapter "Functional divergence on the genomic level", which looks rather humble when it stands alone as it currently does.

We appreciate the interest of the reviewer in this supplementary chapter. To keep the length of the manuscript digestible for a broad set of readers, we decided to only include in the main text the functional differences that could play a role in adaptation to the new laboratory environment.

We consider that a full description of the metabolic differences between the three clades has to be published, as it might be relevant for researchers interested in L. plantarum metabolism. However, it does not fully follow the storyline, as the differences reported in the supplementary, such as nitrate respiration or synthesis of molybdenum cofactors, might not be involved in the clade-specific selection observed in the time series.

(7) Line 773: "Clades C and H encode a shared genetic repertoire related to sugar/riboflavin metabolism that is lacking in clade U". This indeed allows us to hypothesise that the fixation of these clades in fly populations was due to their improved metabolic capabilities. However, the analysis of fitness shows similarity in flies associated with clades H and U, meaning that sugar/riboflavin metabolism in H does not provide an obvious adaptive trait to flies. Moreover, one could say that sugar metabolism in clade C is maladaptive not only for flies, but also for bacteria in liquid cultures. It is recommended to more clearly state the respective limitations of the study.

Here we have to make a distinction between bacterial fitness and host fitness. The three clades differ in their (bacterial) relative fitness, as evidenced by the time-series dynamics (Figure 4). In the cited statement we hypothesize that a more versatile sugar metabolism repertoire could increase the bacterial fitness of clades H and C (relative to clade U) in the sugar-rich laboratory diet.

This is independent of the fitness effect that L. plantarum could have in the host. Finally, as it was discussed in the recommendation 3, fitness is specific to the environment. Clade C is the least fit in liquid MRS in hot conditions, but the fittest in cold experimental conditions.

(8) The authors should better explain why growth in MRS was not performed in a cold temperature regime to further support or refute the hypothesis that capacity and inflection time could partially explain the higher fitness of bacterial strains from clade U.

We did not perform this experiment in cold conditions due to technical limitations of the plate reader, that does not have cooling capacity. Nevertheless, following the reviewers' suggestion, we have included in the revised version of the manuscript a new MRS growth experiment in cold-like conditions (constant 20 °C).

(9) When mentioning that L. plantarum can "increase larval fitness of Drosophila melanogaster relative to germ-free flies" (line 196), the authors should specify in which specific conditions this phenotype was observed, and how relevant the mentioned phenotypes are to the current study.

Following the reviewer's recommendation, we have modified the paragraph in order to clarify the conditions used in other papers and those used in our work. The references cited in this section (PMID: 21907145, 29290388 and 28062579) report that L. plantarum increases the host fitness in protein-poor diets (12 g/l of dried yeast or less), but not in high-protein diet (50 g/l of yeast or higher). Since our experimental diet contains an intermediate amount of protein (24.3 g/l of dried yeast) we were agnostic of whether L. plantarum would benefit the host or not in our conditions. Regarding the phenotypes, we chose two reproductive traits that are affected by changes in the microbiome according to the literature. Developmental time is directly affected by L. plantarum in the aforementioned papers. Offspring number is another fitness component affected by Drosophila microbiome (PMID: 30510004).

(10) Provide a reference for line 205: "In axenic D. melanogaster none of the L. plantarum clades provided a fitness advantage to the host relative to germ-free controls, contrary to the effects reported in the literature". If the conditions were different from those in the studies referred to, then it would be of no use to compare the fitness advantage (for example, in Reference 24 another type of diet was used).

Already covered in recommendation 9.

(11) Please provide more context to this statement (Line 210): "The high content of dried yeast 24.3 g/l in the fly food used in our experiment likely provided already sufficient amounts of essential amino acids, which negated the growth-promoting effects of L. plantarum". It is not clear why amino acids are taken into account, and what the evidence is for the fact that the amount of essential amino acids was sufficient to abolish growth-promoting effects.

The whole paragraph was modified in order to clarify the relationship between protein input and nutritional fitness benefit of L. plantarum.

(12) Please provide measurements of bacterial quantity which would support the statement (Line 215): "The fitness reduction was stronger in the first transfer of flies, likely due to a higher bacterial load".

Upon request of the reviewer, we have estimated the bacterial load per individual fly in D. simulans. Additionally, we have specified the CFUs inoculated in the vials in transfer 1.

(13) Correct the typo (line 220): "However, the developmental time was significantly extended after inoculation with clade C at cold temperature (Dunn's test, p < 0.05 05 for all significant comparisons)".

Done.

(14) Specify more precisely the temperature conditions referred to in line 226: "In summary, we observed that clade C, which is dominant in the cold-evolved populations, decreases host fitness when axenic flies are inoculated". Does it decrease fitness both in hot and cold environments?

For the axenic flies, we did find a decrease in fitness in both regimes, yes. We specified it in the revised version of the manuscript.

(15) Please provide evidence for line 227, or otherwise rephrase it: "The magnitude of this effect varies depending on the environmental temperature, the bacterial load, and the presence of other microbial taxa".

Novel evidence was provided regarding the role of bacterial load on host fitness.

(16) Correct the following statement, so that it reproduces the results of the original work (reference 19, line 229): "In a low-protein diet, strains that were not isolated from Drosophila enhanced larval growth relative to germ-free individuals, whereas another Drosophila-associated strain did not have any effect".

This statement was removed from the revised version. This reference was cited in the discussion to state that: " the nutritional symbiosis in L. plantarum is strain-specific".

(17) Please provide a rationale for using KEGG Orthologs. Why was this database chosen as an appropriate one, even though it is known to be a non-exhaustive metabolomic resource?

KEGG is a well-known metabolic database that is widely used in comparative genomics (PMID: 40177264) and built in state-of-the-art software for microbial ecology such as Anvi'o (PMID: 33349678). Other similar gene-to-function databases are less focused on metabolic pathways, such as COG or GO, or limited to specific enzymatic activities, like CAZy. Furthermore, the hierarchical organization of KEGG Orthologs in modules and pathways allowed us to map clade-specific orthologs to the broad metabolic context. For these reasons, we considered KEGG to be the best option for this analysis.

(18) Line 250: "Therefore, we speculate that the ability to exploit this ubiquitous source of carbon and nitrogen in the lab-maintained fruit flies, could be a strong target of selection in the lab environment". This statement concludes the "Results" section but would be more appropriate for the Discussion section, since the authors do not provide any experimental evidence that could support this statement.

We have modified this chapter, as covered in recommendation 5.

(19) Line 276: "However, the intraspecific richness of L.plantarum in our flies was three times higher than that estimated in human gut microbiomes". Note that there are other recent studies which show the presence of several OTUs within L.plantarum isolates (for example PMID: 41484402).

We thank the reviewer for the reference. We comment on it in the revised manuscript.

(20) Line 287: "Our finding shows that the well-characterized nutritional symbiosis between Drosophila and L. plantarum depends on the bacterial genotype and cannot be generalized to the entire species". Note that such a conclusion has already been previously stated (for example, PMID: 30008290 and 28993620).

We thank the reviewer for the references. Indeed, these papers show that some L. plantarum strains are beneficial for the host while others are neutral. Furthermore, as commented by Reviewer #1 in the public review, L. plantarum has been shown to reduce the host's fitness relative to axenic flies (Gould, PNAS, 2018).

Our observations are novel in two ways. (1) The fecundity observed in D. melanogaster, 0.06 offspring/female/day in average, is lethal (in Gould et al. 2018 fecundity never decreased below 1 offspring/female/day). (2) Clade C outcompetes the other clades in the cold, despite being detrimental for the host.

We have modified the Discussion to account for the previous work.

(21) Line 343: "In addition, we obtained L. plantarum genomes from two other experimental evolution studies. Two genomes from the South African experiment and seven genomes from the Portugal experiment". Merge two sentences into one.

Done.

(22) Line 360: "At sampling, the age of the flies varied between four and eight days for the hot environment and between nine and 16 days for the cold environment". Please comment on the fact that different age of flies (different physiology) is not the reason for bacterial community differences.

During maintenance, flies are sampled at different ages because the temperature affects their developmental time. We cannot rule out the hypothesis that age difference drives microbiome differences. Temperature could affect clade competition directly (differences in optimal temperature between clades) or indirectly, by affecting either the host (e.g. changes in Drosophila developmental time alters L. plantarum fitness), the surounding microbiome, or the food (e.g. increased metabolic activity in the hot regime changes nutrients profile). We ruled out the direct effect of temperature with growth experiments in liquid MRS medium and solid fly food, but disentangling the indirect effects is not feasible.

(23) The majority of figures have low-quality labels that are not legible due to the small size of the font. Please improve.

Done.

(24) Figure 1 - Correct the legend: There is no "10" label on the picture. Probably by 10, the authors mean "Generation", while by x10 - number of isogenic replicates.

Done.

(25) Figure 2 - No numbers at nodes are indicated, whereas it is announced in the legend that they represent bootstrap support values. In addition, it is recommended to show a reference pangenome in the middle panel to clearly refer to the total size of the possible black bar.

We added high bootstrap support as coloured nodes in figures 2, 3 and S2.

We do not understand the reference pangenome request. In the middle panel, each black/white bar corresponds to an orthologous gene that can be either present or absent in each of the genomes. These orthologs were sorted based on hierarchical clustering of the their patterns of abundance (columns present in the same set of genomes, together), not by synteny. Thus, a reference pangenome would be simply a black bar.

(26) Figure 2: It would be advantageous to add a figure that represents the frequency of each strain in each replicate (at the last time point, for instance). It would explain why some “blue” strains appear to be within the “red” cluster. Otherwise, it is confusing to find cold-evolved bacterial strains in hot-evolved fly populations.

The frequency of each clade in each replicate is shown in figure 4. We think that it would be more confusing to follow the suggestion of the reviewer, as the isolates were sampled at different time points of the experiment. We would not like to call it a confusion that "blue" strains appear in the "red" cluster, but rather the logical consequence of the color code used in figure 2, which corresponds to the temperature regime in which the isolate was sampled (regardless of its clade). Whereas in the following figures colour represents the clade. It is thus possible to find clade H isolates in the cold temperature regime, as this clade is in low frequency but not completely absent in this regime.

(27) Figure 3 – Add a label for the X-axis.

We rotated the tree to be able to increase the genome IDs. We have added the label to the Y-axis.

(28) Figure 4 - Please indicate how the clade relative abundance was assessed.

Clade relative abundance was inferred by mapping competitively the short reads against the three clades’ reference sequences. It is specified in the legend now.

(29) Figure 6 - Total number of F1 flies eclosed normalised by day (during which period?). What do T1 and T2 correspond to?

During the respective number of days that females were allowed to lay eggs: one day in the hot settings and two days in the cold settings in transfer 1. One day and three days, respectively, in transfer two.

T1 and T2 correspond to the first and second transfers, as described in the Materials and Methods. In first transfer, flies laid eggs in vials pre-inoculated with a set load of L. plantarum. After egg laying, same adults were then transferred to a sterile set of vials and allowed to lay eggs again (second transfer). Bacterial load in these vials was solely seeded by the parents.

In order to avoid any confusion, in the revised version of the manuscript we have modified figure 6 to show transfer 1 for both Drosophila species, and moved transfer 2 dataset to supplementary figure S5.

(30) Figure S2 - label the X-axis.

We guess the reviewer means Y-axis. Done.

(31) Figure S3 demonstrates the real data and its variability, so it would be better used instead of Figure 4 (which seems to be just a derivative from Figure S3, not a separate dataset and separate type of analysis).

As the reviewer suggested, we have replaced figure 4 with figure S3.

(32) Figure S4: Improve plot title: (e.g., C:H:U = 3:3:3).

Done.

(33) Figure S6: It is stated that N = 10; however, some datasets do not have 10 points represented. Please specify why. Also, please specify the meaning of "T1/T2".

For the inoculation experiment in Drosophila simulans, we had nine replicates per treatment, not ten. This has been corrected in the figure and in the Materials and Methods section.

T1 and T2 correspond to the first and second transfers, already covered in recommendation 29.

(34) Table S3: provide legend for values (1 - present in all strains, but 0.04 - what does it mean?).

It means that 4% of the genomes from this clade harbour the specific gene. We have specified it in the legend of the revised table.

Version 2

Abstract

Most microbiome research focusses on the taxonomic composition at the species level to understand the impact of environmental factors, but intraspecific diversity has largely been ignored. To address this significant knowledge gap, we took advantage of the simple, culturable microbiome of Drosophila. First, we documented that natural populations of D. simulans harbor three diverged clades of Lactiplantibacillus plantarum, a key nutritional symbiont. We studied the distinct ecological roles of these three clades by exposing flies with their native microbiome to two temperature regimes in the laboratory. Tracking the three clades within the complete Drosophila microbiome over a period of more than 10 years at two temperatures, we identified strikingly distinct dynamics in response to the selection regime. We confirmed the functional differentiation of the three clades using in vitro growth measurements and in vivo mono-association assays. Our results highlight that environmental selection operates at the subspecies level. Therefore, we conclude that the functional diversification of the microbiome can only be understood when intra- and interspecific diversity is considered.

Introduction

Intraspecific diversity has largely been overlooked in microbial ecology surveys. This can be in part attributed to the success of 16S rRNA profiling, which allows the characterization of species diversity with a very high efficiency (1,2), but cannot identify intraspecific diversity (3). Although environmental variation has been identified as one of the major factors driving species composition, its role for intraspecific adaptation has been largely overlooked (3). This blind spot of microbiome research is particularly surprising as it is known that conspecific strains, i.e. strains from the same species, can display large phenotypic variation (4). Clinically relevant species like Escherichia coli, in which different subspecies can be either pathogenic or commensal are a particularly illustrative example (5).

Although a few studies in environmental microbial communities have provided evidence for the functional importance of intraspecific variation (6,7), such analyses are considerably more difficult. Therefore, descriptive correlative studies are more common than functional assays (8). In the gut-dwelling bacterium B. fragilis conspecific strains compete for the same niche leading to the mutual exclusion of toxicogenic and non-toxicogenic strains in the gut environment (9). Not only in the human gut, but also in lakes, soil, and the sea conspecific strains frequently co-occur in sympatry (10–12). It has even been even reported that strain richness, i.e. sympatric strain diversity within the same species, stabilizes microbial communities in fluctuating environments (13,14).

In this work, we further explored the functional implications of intraspecific diversity within the same microbial population. We focused on the species Lactiplantibacillus plantarum, an extremely versatile lactic acid bacterium that has been associated with a variety of different habitats, including plants, the gastro-intestinal tracts of mammals and insects, as well as food such as meat, dairy, and pickled products (15). L. plantarum is a facultative symbiont of Drosophila (16–19). Drosophila larvae can enhance the growth of L. plantarum, which in response releases essential nutrients for the larvae under poor diet conditions (18,20). As a result of this nutritional symbiosis, L. plantarum is a prevalent taxon in the Drosophila microbiome (17).

To study the implications of intraspecific diversity of L. plantarum in fruit flies, we took advantage of an experimentally evolved Drosophila simulans population that has been adapting to two novel temperature regimes, hot and cold, for more than ten years. We detected three clades of L. plantarum co-occurring within the native microbiome of D. simulans population before the experimental temperature was modified. However, the exposure of the fly population to the novel temperature regimes altered the intraspecific composition of L. plantarum, revealing functional differences between the co-occurring clades. These functional differences were experimentally validated in vitro and in vivo. Our results provide a clear example for functional diversification of intraspecific diversity and demonstrate how experimental evolution can uncover this otherwise hidden functional diversity.

Results

Genome sequencing reveals the presence of three L. plantarum clades in the experimentally evolving populations

We sampled 33 isolates from the Florida experiment (See Materials and Methods for details) to characterize the diversity of L. plantarum (Figure 1, Table S1). The isolates formed three clades based on the pairwise average nucleotide identity (ANI). Between-group ANI values were 98.4-99.17%, below the ∼99.5% threshold for conspecific strain identity (21). On the other hand, the identity within groups was 99.48-99.99% (Figure 2). The clustering was consistent irrespective of whether we used SNPs in the core genome or presence/absence patterns of accessory genes. This indicates that the three clades are old with very limited genetic exchange.

Experimental setup to study the long-term evolution of L. plantarum in Drosophila simulans populations.

Pangenome of L. plantarum genomes isolated from the Florida experiment (Table S1), sorted by phylogeny.

Left panel: Maximum likelihood tree built based on the core genome alignment. Tree leaves correspond to the name of the isolate and are coloured based on isolation origin: ancestral population (grey), hot-evolved population (red) or cold-evolved population (blue). Nodes supported by bootstrap values ≥ 90% based on 1000 replicates are highlighted in yellow. Middle panel: Pattern of genes presence (black)/absence (white) in each genome. Right panel: pairwise average nucleotide identity between the isolates. The colour ranges between 98.5% (purple) and 100% (yellow).

Since the isolates were sampled either from the unevolved, hot-evolved or cold-evolved populations (Figure 1), we tested whether the clades were randomly distributed across the experimental treatments and observed a non-random distribution (Fisher’s exact test, 3x3 contingency table, p = 5.5e-10). In combination with the clustering of clades (Figure 2), we conclude that each of the three clades is more prevalent either in the novel experimental populations or in the unevolved founder population, at the beginning of the experiment. We therefore coined the terms H (hot-evolved), C (cold-evolved), and U (unevolved) clades, that will be used throughout the manuscript.

In order to further investigate the association between L. plantarum genotype and temperature, we sampled from two additional experiments, obtaining seven isolates from hot-evolved Drosophila simulans Portugal and two isolates from constant-cold-evolved Drosophila simulans South Africa (Table S1). In a phylogenetic tree, the hot-evolved isolates and the constant-cold-evolved isolates grouped into clades H and C, respectively (Figure S1). These results do not only support the association between L. plantarum clade composition and environmental temperature (Fisher’s exact test, 3x3 contingency table, p = 2.6e-12), but also suggest that clade C is favored irrespective of whether the temperature is fluctuating (20 °C during the day, 10 °C at night) or constant (15 °C). Furthermore, the grouping of isolates originating from different experiments to the same clades indicates that these are not restricted to a single natural D. simulans population, but represent global diversity.

The three sympatric clades independently adapted to their host

Our analyses suggest that all three clades are shared among natural D. simulans populations, but it is not clear whether they diverged in D. simulans or independently colonized their Drosophila host. We used 79 publicly available L. plantarum genomes from several environments and geographic locations (Table S2) to reconstruct the phylogeny based on the core genome.

All three L. plantarum clades are more closely-related to genomes from other sources than to each other, which implies that they diverged before colonizing D. simulans (Figure 3). Moreover, the genomes from the clade H are almost identical to public sequences isolated from global collections of Drosophila melanogaster (Figure 3, Table S2). The clade U isolate B246 is closely related to the Drosophila-associated isolate LpWF originating from a wild D. melanogaster individual (22). Clade C did not have a close relative in the collection. The high similarity of the L. plantarum sequences isolated from D. melanogaster and D. simulans highlights that both Drosophila species share this component of the microbiome. Most likely a deeper sampling of the D. melanogaster microbiome will also detect clade C.

Maximum likelihood tree of 92 L. plantarum genomes.

Tree leaves are coloured by source of isolation. The genomes from our lab are coloured according to experimental treatment: unevolved in grey, hot-evolved in red, and cold-evolved in blue. Nodes supported by bootstrap values based on 1000 replicates are highlighted in orange (≥ 70% bootstrap support) or yellow (≥ 90% bootstrap support).

Despite being phylogenetically divergent, the three sympatric clades could have converged functionally in the process of adaptation to a common environment. In order to test this, we built the L. plantarum pangenome including all the available isolates and performed hierarchical clustering based on the presence/absence of accessory genes. We found no grouping of the three sympatric clades (Figure S2), which suggests that they are not only phylogenetically diverged, but have the potential to be functionally different based on their pool of accessory genes.

Clade dynamics are temperature-specific and consistent between population replicates

Up to now, our analyses were restricted to isolates sampled at specific time points of the experiment. Since genomic DNA from flies with their microbiome was sequenced throughout the entire experiment (Figure 1), a metagenomic analysis could track the frequency trajectories of the three clades to shed more light on the adaptation process. We calculated the proportion of reads assigned to each clade in the population at a given time point. At the beginning of the experiment, the populations were dominated by clade U (54.5% mean relative abundance), followed by clade C (31.9%), and then clade H (13.6%). When the fly populations were subjected to the experimental treatments, clade H became dominant under the hot regime and clade C became dominant under the cold regime (Figure 4). Clade U greatly decreased in relative abundance in both regimes. This suggests that changes in environmental temperature alter the fitness landscape of the population and lead to a new adaptive state, in which the relative fitness of the clades change.

Clade composition over time in each replicate population from both temperature regimes.

C in light blue, clade H in red and clade U in grey. Clade relative abundance was inferred by mapping competitively short reads against the three clades’ reference sequences.

A strong asset of the experimental evolution set up is the availability of 10 replicate populations that were generated from the same founder population and independently maintained throughout the entire experiment under the same conditions. Hence, the comparison of replicates provides an estimate for the strength of deterministic and stochastic forces during the experiment. In the cold regime, all ten replicates exhibited a rapid takeover of clade C (Figure 4). In contrast, the hot-evolved replicates were more variable in the rate at which clade H increased in abundance. The most extreme cases were replicates 8 and 10, in which clade C dominated the population for 90 generations before being replaced by clade H. Thus, the populations subjected to the hot regime took longer to reach a new equilibrium, which is also less reproducible than that of the cold regime. Overall, the temperature-specific dynamic is consistent across replicates in both regimes, confirming the association between environmental temperature and clade frequency that was observed during the isolation of the clones (Figures 1, 2 and S1).

Clades differ significantly in their growth kinetics in liquid culture

The metagenomic analysis showed that a different clade dominated each novel temperature regime and this was highly consistent across replicates. The temperature-specific dynamics can arise from clade-specific growth kinetics, as observed for B. cereus groups (23). Alternatively, temperature could affect L. plantarum indirectly, by acting on the host, the food, or other microbiome members. To determine if the observed clade dynamics were driven solely by differences in optimal growth temperature, we grew isolates from the three clades in liquid MRS medium at temperature conditions similar to the experimental settings (See Materials and Methods for details). The selected isolates represented all sampling events across the two temperature regimes (hot and cold) as well as the unevolved population (Table S1).

The three clades differed significantly in growth rate, carrying capacity, and inflection time, i.e., the time it takes an isolate to reach the log phase, in both temperature conditions (Figures 5 and S4; Kruskal-Wallis test, p < 0.005 for all parameters). However, the performance of the three clades did not fully match our expectations, based on the time series. In hot conditions, clade U outperformed clade C in growth rate, carrying capacity, and inflection time (Figure 5) despite clade C prevailed in the Drosophila populations for many fly generations (Figure 4). In cold conditions, clade C had significantly lower growth rate than clade H and no significant differences in carrying capacity and inflection time. These results reveal differences in growth kinetics of the three clades, but do not support the hypothesis that differences in optimal growth temperature caused the shifts in composition observed in the time-series experiments in flies.

Clade-specific growth dynamics at hot fluctuating 28/23 °C.

(A) Overlapped curves of all replicates. Lines are coloured by clade. Background colour represents the growth temperature at a given time. (B-D) Boxplots depicting the carrying capacity, growth rate, and inflection time of each clade. We measured the growth of four isolates from clade U, nine isolates from clade C and sixteen isolates from clade H (Supplementary table S1). Each isolate was grown three time. Each dot corresponds to a technical replicate. Data is grouped and coloured by clade. Statistical significance was determined using Dunn’s test with Holm-adjusted p-values. Only significant comparisons are indicated. **** p < 0.0001; *** p < 0.001; ** p < 0.01; * p < 0.05

Nevertheless, we caution that the environment in the time-series experiment is radically different to that of liquid MRS (24). We confirmed this by agar-drop assays in solid fly food, which resulted in delayed growth of clade U, relative to clades H and C, in both focal temperature regimes (Figure S12). Furthermore, L. plantarum lives in a complex community in which the different clades interact with each other, with other members of the microbiome, and finally also with the host. Bacteria can persist in the community by thriving in the food or by stably colonizing the fly gut. The former strategy might be controlled by growth rate and carrying capacity, but the latter is affected by additional factors such as death rate in the gut and defecation rate, that cannot be assessed in liquid medium (22). Thus, observed differences in growth could indicate different strategies to persist in the community.

Fitness effects of L. plantarum strains on their host

Although growth dynamics differ significantly among the three clades in growth culture, this does not explain the observed changes in abundance during experimental evolution. Therefore, we turned to the Drosophila host to explore the fitness consequences of colonization by each clade.

L. plantarum can increase larval fitness of Drosophila melanogaster relative to germ-free flies under low protein conditions (18,20,25). Protein content is controlled by the amount of yeast in the fly food (18,20,25). L. plantarum-mediated benefits have been observed in low yeast (<=12 g/l) food, but not in protein-rich diets (50-80 g/l yeast). Because our experimental diet contains an intermediate yeast content (24.3 g/l), we wondered whether we could detect temperature-dependent fitness effects of the host in this food. To test this, we designed a set of inoculation experiments in both focal temperatures using two host species, D. melanogaster and D. simulans, and the three L. plantarum clades. While we could produce axenic D. melanogaster, attempts to produce axenic D. simulans were unsuccessful. Thus, D. simulans used in these experiments were dechorionated but not axenic.

Contrary to experiments with protein-poor diets, none of the L. plantarum clades provided a fitness advantage to the host relative to controls in any of the species. In the case of clades U and H, the number of offspring and developmental time did not differ significantly from the dechorionated controls (Figure 6; Dunn’s test, p > 0.05 for all pairs). Surprisingly, in both host species we observed significant fitness reductions in the presence of clade C. While in D. melanogaster offspring number decreased and developmental time increased at both temperatures (Figure 6; Dunn’s test, p < 0.05 for all significant comparisons), in D. simulans only developmental time increased in the cold only (Figure 6; Dunn’s test, p < 0.05 for all significant comparisons). We attribute the weaker effects in the non-axenic flies to a reduced L. plantarum load due to the presence of other taxa (22), or to higher-order interactions with other members of the microbiome (26–28).

Fitness effect of L. plantarum inoculation in axenic D. melanogaster and conventionally reared D. simulans in the first transfer (See Materials and methods).

(A, B) Total number of F1 flies eclosed normalized by day and female under the cold (A) and hot (B) regime. (C, D) Developmental time, estimated as the number of days it takes 50% of the offspring to eclose. Measurements were grouped by inoculation treatment and Drosophila species. Each dot corresponds to a biological replicate (n = 10 for D. melanogaster and n = 9 for D. simulans). Statistical significance was determined using Dunn’s test with Holm-adjusted p-values. Only significant comparisons are indicated. *** p < 0.001; ** p < 0.01; * p < 0.05.

We further explored the role of bacterial load on the host fitness by extending the experiment to a second transfer: the same parents laid eggs into a new set of sterile vials (See Materials and Methods for details). The effect of clade C on the host disappeared for D. simulans and was diminished in D. melanogaster relative to the first transfer (Figure S5; Dunn’s test, p < 0.05 for all significant comparisons). Since the bacterial load could be a major factor influencing these results, we performed an additional experiment in D. simulans to quantify the bacterial load of parental flies in the second transfer. Median bacterial load of flies inoculated with clade C was 4.1×105 colony forming units (CFUs) per fly in the hot regime and 1.3×104 CFUs/fly in the cold one, at least 10-fold higher relative to the other clades in the same regime (Figure S6). These results suggest that effect of clade C on the host depends on the bacterial load, and could be related to an excess of bacterial growth inside the host.

In summary, L. plantarum does not confer a fitness benefit to the host in any temperature regime. Moreover, clade C, which is dominant in the cold-evolved populations (Figure 4), is clearly detrimental for reproduction and development in D. melanogaster hosts. In D. simulans, though, slower development provided by clade C could be beneficial in the cold, as observed in other environmental conditions (29,30).

Functional divergence on the genomic level

Our analyses clearly indicate functional divergence between the three clades at genetic and phenotypic levels (Figures S2, 5 and 6). Therefore, we were interested in identifying the genes that contribute to this functional differentiation. A comparative KEGG pathway analysis showed that clade C contains the most KEGG Orthologs (KOs), 1163 KOs, followed by clade H (1157 KOs), and clade U (1131 KOs). A subset of 111 orthologs is unique to one or two clades. Of these, 41 are shared by clades H and C (Figure S7). These differential orthologs fell into 53 KEGG metabolic pathways.

Interestingly, clades C and H encode a shared genetic repertoire related to sugar metabolism that is absent in clade U, which does also not encode any private sugar-related function (Table S3). Clade U cannot use sorbitol as carbon source, since it does not encode the PTS transporter to internalize it (EC: 2.7.1.198; Figure S8) nor the enzyme sorbitol dehydrogenase (EC: 1.1.1.140), that converts sorbitol into fructose (31). Clade U also lacks transaldolase activity (EC: 2.2.1.2), that connects the Embden–Meyerhof–Parnass pathway with the non-oxidative pentose phosphate pathway (32). Finally, clade U is unable to hydrolyze chitobiose into N-acetyl-glucosamine monomers because it lacks the enzyme hexosaminidase (EC: 3.2.1.52).

Additionally, clade H and C harbor different sugar-related functions (Table S3), indicating that, despite partial overlap, these strains possess distinct sugar metabolic capacities. We also identified functional differences between clades in cofactor biosynthesis and respiratory metabolism. These differences do not explain the clade-specific selection, but reflect the different evolutionary histories of the clades (Table S3, figures S8-S11). A detailed description of the functional differences between clades can be found in the supplement.

To sum up, among the clade-specific genes we found an enrichment of sugar-related functions in clades C and H. This, summed to the delayed growth of clade U in solid fly food (Figure S12), suggests that diet, in addition to temperature, affected the intraspecific dynamics of L. plantarum.

Discussion

Intraspecific diversity within the same environment has been reported to be ecologically important to maintain the stability of the entire microbial community (11,13,14). This parameter has been studied in lakes, soils, seas, and the human gut (10–12). However, little attention has been given to intraspecific variation in Drosophila, and insect microbiomes in general (8). Based on our results in experimentally evolved fruit flies, we propose that within-species competition, thus far largely overlooked, could contribute to ecological adaptation and evolution of the host, as it has been observed at the species level in Drosophila (33) and other animals (34–36).

Our finding of three co-occurring subspecies of L. plantarum in our populations is particularly surprising, given the low microbiome richness of Drosophila at species level, which is estimated to be orders of magnitude lower than in humans (37,38). However, the intraspecific richness of L. plantarum in our flies was three times higher than that estimated in human gut microbiomes (11). We note that a recent study also found several co-occurring L. plantarum variants within a Drosophila population (39). Our work further expands this observation by demonstrating that this diversity is not neutral, but rather driven by external factors. This discrepancy between inter- and intra-species diversity suggests that the Drosophila microbiome is more complex than previously thought (16,26,40). However, this diversity exists below the species level and has therefore remained cryptic due to the limitations of amplicon sequencing (3).

By altering the experimental temperature at which the populations were maintained, we discovered consistent shifts in the composition of L. plantarum subspecies, that inevitably led to the dominance of a single clade. Further in vivo experiments revealed that clade composition significantly impacts the fitness of the host. It was previously shown that nutritional symbiosis in L. plantarum is strain-specific (19), and that some genotypes can be detrimental for the host (41,42). In our study, we also showed that clade C was detrimental to the host, but nevertheless in the Drosophila time series experiment, this clade outcompeted the other clades, which did not have a negative effect on host fitness. This demonstrates that the benefits of parasitism or mutualism in the Drosophila microbiome are context-dependent, as already observed in humans (43,44).

Finally, we also gained insight into how the microbiome of the fly could have adapted to laboratory conditions. The two clades, C and H, which are favored in the time-series, harbor unique sugar-related metabolic functions, consistent with faster growth in the sugar-rich laboratory diet. Besides the novel temperature regimes, introduction to the laboratory changes the diet and mobility of the host and the microbiome relative to the wild (45,46). This explains the rapid decrease of clade U in both temperature regimes, despite showing high fitness in liquid MRS medium. Our findings are consistent with microbiome laboratory adaptation, and agree with previous observations in L. plantarum (17) and Acetobacter (46).

Overall, our work emphasizes the importance of intraspecific diversity in maintaining the stability of microbial communities and the host’s ability to adapt to environmental changes. The current climate crisis has sparked a significant interest in how environmental temperature affects microbial communities at the species level (47–49). Our study highlights that even subspecies diversity plays a key role in adaptation to environmental temperature, and probably other abiotic factors. Therefore, we propose that a comprehensive understanding of adaptive processes and ecological dynamics in the case of host-microbiome interactions depends critically on including diversity at the subspecies level.

Materials and Methods

Experimental evolution

Twenty replicate populations were set up using 202 isofemale lines from a natural D. simulans population collected in Florida (50). The populations were maintained in a 12 h photoperiod and at two different novel temperature regimes: ten replicates were kept in a fluctuating hot environment (28 °C during the day and 18 °C at night), whereas the other ten were kept in a fluctuating cold environment (20 °C during the day and 10 °C at night). The census population size of the replicates was 1000-1250 with a 50:50 sex ratio. The flies in each replicate were equally distributed across five 300 ml bottles containing 60 ml of standard Drosophila medium (300 g Agar + 990 g sugar beet syrup + 1000 g malt syrup + 2,310 g corn flour + 390 g soy flour + 900 g yeast in 37.5 L water) (51). The evolved D. simulans populations were sequenced in 10 generation intervals using Pool-Seq (52,53). A subset of these sequencing reads has been used to study the evolutionary dynamics of fruit flies (53). Similar pool-seq time series data have been previously used to analyze long-term endosymbiont and microbiome dynamics in D. melanogaster populations evolving in the same experimental conditions (54–56).

We also isolated L. plantarum from two other experimental evolution experiments, one originated from a South African population reared at constant 15 °C and the other from Portugal, which was reared in the fluctuating hot regime as described earlier (57). Flies in both experimental evolution studies followed the same maintenance regime as the Florida populations, with temperature being the only experimentally altered variable (58).

Bacterial isolation

We primarily obtained data from the experimental evolution studies using the Florida D. simulans founder population and collected 33 L. plantarum genomes from five different time points; two time points from each experimental temperature and one time point corresponding to the ancestral flies (Table S1). In addition, we obtained two and seven genomes from experimental evolution studies originating from South African and Portuguese D. simulans populations.

Around 100 flies from each population were homogenized in sterile PBS using an autoclaved pestle. The homogenate was diluted to avoid physical contact between colonies and streaked on agar plates with three different media: De Man, Rogosa, and Sharpe (MRS), mannitol and tryptic soy. The plates were incubated at 28 °C for 48 h. We identified L. plantarum using PCR with custom Lactiplantibacillus-specific primers (in 5’-3’ orientation, LacF: GATGGGCGCTTACCCGATTA, LacR: CTGCCCCGCAAATTGTTTCA). Colonies which could successfully be amplified were picked and regrown to stationary phase in the same medium from which they were isolated, but in liquid format. A glycerol stock of each isolate was made by mixing the culture with glycerol (50% v/v) in 1:1 proportion. The proportion of L. plantarum colonies, relative to those of other taxa, varied across growth media and fly populations. A list of other microbial taxa isolated from the flies can be found in Table S4.

DNA extraction and sequencing

Genomic DNA was extracted from all the replicate populations using the high salt method (59). The fly populations were sampled in 10 generations intervals starting from generation 0. At sampling the age of the flies varied between four and eight days for the hot environment and between nine and 16 days for the cold environment. Pools of flies were sequenced at various time points, using a range of library kits, insert sizes, and read lengths (Table S5). We expanded the genomic dataset used in Barghi et al. 2019 (53) with a new set of pools covering the hot-evolved populations up to generation 250 and the cold-evolved population until generation 100.

For sequencing, the L. plantarum isolates were grown to stationary phase in MRS medium. gDNA was isolated with the high salt extraction method (59). Additionally, a lysozyme pre-treatment was used to degrade Gram-positive cell wall (60). Briefly, the pellet was resuspended in 480 µl of EDTA 50 mM. The suspension was treated with 120 µl of lysozyme (20 mg/ml dissolved in NaCl 30 mM – EDTA 2mM) and incubated for two hours at 37 °C. After this, the samples were centrifuged at maximum speed, the supernatant was discarded, and the pellet was further treated following the high salt gDNA extraction. Genomic DNA was quality-controlled on an Agilent Bioanalyzer (Agilent Technologies, Inc., Santa Clara, CA) and subsequently used to prepare DNBSEQ short-read libraries, and 2x150bp reads were sequenced on a DNBSEQ-G400 (MGI Tech Co., Ltd., Shenzhen, China).

Genomic assemblies

Sequencing reads were processed using BBDuk (BBMap, 2025 v38.90) to remove adapters, ΦX174 sequences and low-quality bases with the following parameters: ktrim=r, k=23, mink=11, hdist=1, tbo, tpe. Reads were assembled using SPAdes v4 (62) with kmer sizes 21, 33, 55, 77, 99, and 127. Subsequently, CheckM2 v1.0.1 (63) was used to estimate the completeness and contamination of each assembly, and the taxonomy was determined with GTDB-Tk v2.1.1 (64). Ten assemblies had more than 5% contamination, probably due to accidentally picking two adjacent colonies. We used Centrifuge v1.0.4 (65) to classify and exclude contigs with exogenous origin from these assemblies.

Comparative genomics and phylogeny

We built the pangenome and phylogenetic tree of the collection of genomes isolated from our lab (Table S1). First, we used Prokka v1.14.6 (66) to predict the genes in all the genomes. Roary v3.13 (67) was used to construct the isolates’ pangenome and extract the aligned genes belonging to the core genome, i.e. those that are present in all genomes. SNP-sites v2.5.1 (68) was used to extract the variable positions from the concatenated core genes. A maximum likelihood tree of the aforementioned variable positions was built using the model GTR+ASC and 1000 bootstraps with IQ-TREE v3.0.1 (69). Additionally, pairwise average nucleotide identity between all the isolates was computed using the program PyANI v0.2.13.1 (70), with ANIb as alignment method using the whole genomes. This workflow was carried out for the subset of 33 genomes isolated from the Florida experiment and for the whole dataset of 42 genomes, including those isolated from South Africa and Portugal experiments. Unless specified otherwise, subsequent analyses were conducted using the whole dataset pangenome.

The same phylogenetic and pangenomic methods were used for the extended L. plantarum genomes collection, including 79 publicly available genomes (Table S2). In order to root the tree, an additional tree was made including the reference genome from the sister species Lactiplantibacillus paraplantarum (GCA_003641145.1) as outgroup. The L. plantarum isolate that was most closely related to L. paraplantarum, NIZO2776, was used to root the tree.

In order to investigate which metabolic functions were differentially encoded in each clade, we followed the Anvi’o v7.1 pangenomic workflow (71,72) to identify “gene clusters” that are unique or shared across the genomes isolated in our lab. These gene clusters were further annotated against the KEGG database using the built-in Anvi’o function ‘anvi-run-kegg-kofams’. The script ‘anvi-compute-functional-enrichment’ (73) was then used to screen for KEGG functions that are differentially enriched between the clades. Briefly, this script estimates the fraction of genomes from each clade that encode each KEGG Ortholog (KO) and tests whether they are significantly associated to one or more groups. We used the R package ggKegg v1.4.1 (74) to visualize the KEGG pathways in which at least one gene was enriched.

Competitive mapping and clade relative abundance calculation

First, we pre-filtered the Drosophila pool-seq reads by removing the reads that have eukaryotic or Wolbachia origin. We used Bowtie v2.5.4 (75) with stringent settings (-D 500 -R 40 -N 0 -L 20 -I S,1,0.50 --no-mixed --no-discordant) to map each pool-seq data set against a collection of reference genomes that included D. simulans (GCF_016746395.2), D. melanogaster (GCF_000001215.4), D. mauritiana (GCF_004382145.1), H. sapiens (GRCh38), M. musculus (GCF_000001635.27), A. thaliana (GCF_000001735.4), S. cerevisiae (GCF_000146045.2), C. lupus (GCF_011100685.1) and a several insect-infecting Wolbachia strains (GCF_000008025.1, GCF_000022285.1, GCF_000376585.1, GCF_000376605.1, GCF_000475015.1). We kept the read pairs in which none of the reads mapped to any of the genomes in the collection, which should predominantly represent prokaryotic sequences.

Then, we used SuperPang v0.9.6 (76) to build the reference graph pangenome of each L. plantarum clade using all the genomes belonging to the clade as input (Table S1, Figure 2). In order to calculate the clades relative abundance in each pre-filtered pool-seq, we used BBMap (BBMap, 2025 v38.90) with “perfect mode” settings to competitively map each read against the three concatenated clade pangenomes. For each sample, the number of reads that mapped uniquely to each clade was counted and normalized to Reads Per Million (RPM) in order to account for differences in genome size and sample depth. L. plantarum represented between 0.0438% and 17.2246% of relative abundance of the reads in all time points.

We benchmarked our approach to confirm the relative abundance estimates from our competitive mapping pipeline. Briefly, we simulated 200k HiSeq reads from each reference graph pangenome using InSilicoSeq v2.0.1 (77), and used ‘seqkit sampl’ v2.3.0 (78) to mix them in different ratios, maintaining the original size of 200k reads: 1:0:0, 0:1:0, 0:0:1, 1:1:1, 2:1:7, and 8:1:1. Each of the combined read sets was randomly subsampled from 150k to 10 reads (10, 20, 40, 60, 80, 100, 200, 300, 400, 500, 600, 700, 800, 900, 1k, 2k, 5k, 10k, 150k reads) and mapped competitively to the three clades. We confirmed that the relative abundances were correctly estimated with as few as 100-1000 reads mapping the references (Figure S3).

Growth assays

The glycerol stocks from all the available L. plantarum isolates were re-grown in agar MRS. A single colony per isolate was grown in liquid MRS and passaged daily for five days in order to remove the influence of freezing. On the fifth day, each suspension was normalized to an optical density (OD600) of 0.01, and serially diluted 1:100 in MRS. We selected 31 isolates that represented samples from all the available time points and temperature regimes, and plated them in triplicates in a flat-bottom 96-wells plate. Three wells were filled with sterile MRS medium as blank controls. Later, we detected contamination in two of the isolates, leaving us with 29 isolates. In total we grew four isolates from clade U, nine isolates from clade C and sixteen isolates from clade H (Supplementary table S1). The plate was incubated for 72 h and OD600 was measured every 15 minutes in a Synergy H1 plate reader. The temperature regime was chosen to be similar to the hot conditions of the fly populations, but due to limitations of the spectrophotometer’s cooling capacity instead of 28 °C during the day and 18 °C at night, we set it to 28 °C during the day and 23 °C at night. Orbital shaking was set to 425 cpm.

Growth in cold conditions could not be measured with the plate reader as it does not have cooling capacity and the experiment would have lasted more than a week, so we used an alternative approach. We grew twelve isolates, four from each clade (Table S1), in glass test tubes using 5 ml MRS. We normalized each suspension to an OD600 of 0.05 and then diluted 1:100 in 5 ml of MRS. We also prepared an MRS blank. We prepared six replicates per isolate and blank. Three replicates were set at 17:00 and three the next day at 10:00. All tubes were incubated at constant 20 °C and 168 rpm in a shaker (Excella E24, New Brunswick, USA). The OD of all tubes was measured for twelve days at 10:00 and at 17:00 using a spectrophotometer (Ultrospec 10; Biochrom, Cambridge, UK). By using the replicates at two different days, we effectively duplicated the measured time points without the need to increase the sampling frequency.

The R package GrowthCurver v0.3.1 (79) was used to infer the growth parameters from the growth curves. All statistical analyses were conducted in R v4.4.2 (80) using the package rstatix v0.7.2 (81). We used the Kruskal-Wallis test to assess overall differences in growth parameters among the clades. For post-hoc pairwise comparisons, Dunn’s test was applied, with p-value adjustments based on the Holm method. Differences were considered significant at p < 0.05.

Agar-drop assays

One representative isolate from each clade (Table S1, B89 for ancestral, S103 for cold, and S239 for hot) was grown overnight in MRS medium from glycerol stocks. OD600 was normalized to 0.2, and subsequently diluted 1:10 six times. We plated on fly food Petri dishes 10 µl drops of the original suspension and of 10-2, 10-4, and 10-6 dilutions. We incubated three plates at 28/18 °C and three at 20/10 °C for 10 days. Pictures of all plates were taken at incubation days 3, 6 and 10.

Production of axenic flies

We generated germ-free flies, using a modified version of the dechorionation protocol described by Kietz et al. (82). Specifically, we used 2.8% active chlorine bleach instead of 1% and supplemented bleach, EtOH, and deionized H2O with TritonX (1% v/v). The addition of this detergent prevents dechorionated eggs from sticking to the walls of the tubes, thus facilitating their transfer to bottles. The whole process was carried out in sterile conditions using a laminar flow hood. Axenic flies were maintained via periodic transfers to new sterile food inside the laminar flow hood. We first carried out this procedure with several D. simulans isofemale lines, but it was not possible to fully eliminate the residual microbiome despite several rounds of treatment. However, we could produce and maintain axenic D. melanogaster Oregon-R flies. For this reason, we decided to use D. melanogaster as axenic host in the inoculation experiments.

Inoculation experiments and measurement of bacterial load

The representative isolate from each clade (Table S1, B89 for ancestral, S103 for cold, and S239 for hot) was re-grown from the glycerol stocks in MRS medium. After three transfers, OD600 was normalized to 0.05. Autoclaved vials with axenic fly food were inoculated with 50 µl of each bacterial suspension. This volume corresponds to ∼2.5×105 CFUs based on colony counting of serial dilutions. In total five treatments were used: the pure culture from each genotype, a mixture of the three genotypes in equal proportions (“mix”), and control, in which the vials were inoculated with 50 µl of fresh MRS medium. After overnight absorption of the liquid, twenty axenic flies (ten female and ten male flies) were added to each vial and allowed to lay eggs for either 24 h (hot conditions) or 48 h (cold conditions). After this step, the flies were transferred to a new set of vials with axenic, uninoculated food, and allowed to lay eggs for another 24 h (hot conditions) or 72 h (cold conditions). The adults were discarded and the vials were incubated at hot (28 °C during the day and 18 °C at night) or cold conditions (20 °C during the day and 10 °C at night) until the new generation eclosed. The number of eclosed flies was recorded daily. We estimated the total number of flies per vial and the developmental time as proxies of fitness. We calculated developmental time as the day from the start of the experiment until the day at which 50% of the flies in the vial were eclosed. The inoculation experiment was carried out with axenic D. melanogaster Oregon-R (ten replicates per treatment) and dechorionated, but non-axenic, D. simulans from Florida (nine replicates per treatment).

To quantify bacterial load of individual flies, we transferred four D. simulans males to two sets of inoculated vials as in the previous experiment. We allowed the flies to feed for 24 h in either hot or cold conditions before transfer to sterile food and maintained in the same regime for 72 h. Each fly was then placed in a 2 ml tube with 200 µl of PBS and a sterile 4 mm glass bead. The tubes were shaken at 1450 rpm for 4 minutes in a MiniG 1600 vertical shaker (SPEX®SamplePrep). Homogenates were diluted 1:10 (cold-treated) or 1:100 (hot-treated) in PBS, and 200 µl of each dilution was plated in an MRS plate. After overnight incubation at 28 °C, CFUs were counted in each plate. Based on colony morphology, most CFUs corresponded to L. plantarum in inoculated flies, despite using D. simulans harboring microbiome.

We used the Kruskal-Wallis test to assess overall differences among the treatments. For post-hoc pairwise comparisons, Dunn’s test was applied, with p-values were adjusted using the Holm method. Differences were considered significant at p < 0.05.